### 2023-07-12 R/least-cost-paths.qmd
🚀 initial commit

---
title: "Supplement A: Estimating Travel Cost"
date: today
format:
  html: 
    embed-resources: true
    fig-align: center
    fig-dpi: 300
    fig-responsive: true
    number-sections: true
    number-depth: 3
    toc: true
    toc-depth: 3
    smooth-scroll: true
    theme: cosmo
    highlight-style: a11y
execute: 
  echo: true
  warning: false
  error: false
---

<!-- 
RStudio/Quarto is having some issues with figure sizing, 
so using this hacky CSS solution. 
-->

```{css}
#| echo: false

.img-45 {
  width: 4.5in;
}

.img-65 {
  width: 6.5in;
}

```

## Overview

**Goal:** estimate cost-distance (in time) from each ranch to Ciudad Constitucion, La Paz, and local springs.

**Data:** a Digital Elevation Model (DEM) along with point locations for ranche clusters, springs, cities.

**Method:** Campbell's hiking function [@campbell2019].

There are two watersheds in the project area. Each one has a single road that connects it to Highway 1 and thus to the markets in Ciudad Constitucion to the north and the capitol La Paz to the south. The "road" that connects the watersheds to each other is not maintained and thus rarely used. As a consequence, nearly all of the variance in travel time to cities is found within each watershed. For that reason, we estimate the time it takes to leave each watershed ("leaving" here means hitting the edge of the project area), rather than the true time it takes to get to each city. We interpret this as the relative difference in time it takes to get to Ciudad Constitucion from the northern watershed and the relative difference in time it takes to get to La Paz from the southern watershed. To get to La Paz from the northern watershed, we simply add an arbitrary half hour to the total time to get out of that watershed. The same would presumably be the case to get to Ciudad Constitucion from the southern watershed, though that value does not factor into our analysis.

For springs, Dr. MacFarlan's ethnographic data suggest that individuals residing within the same ranch cluster tend to rely on one or two springs at most. Thus, we estimate road distance to the two nearest springs and take the average.

All spatial data are projected to the Mexico ITRF92 / UTM zone 12N CRS ([EPSG:4485](epsg.io/4485)).

## R Preamble

```{r libraries}

library(dplyr)
library(ggnewscale)
library(ggplot2)
library(here)
library(hiker)
library(sf)
library(terra)
library(viridis)

```

Specify consistent plot theme here:

```{r}
#| code-fold: true

# this is designed to position the legend 
# at the top left corner of the plot area

theme_set(theme_bw(12))

theme_update(
  axis.text = element_text(size = rel(0.5), color = "gray45"),
  axis.text.y = element_text(angle = 90, hjust = 0.5),
  axis.ticks = element_line(color = "gray45", linewidth = 0.1),
  axis.ticks.length = grid::unit(0.1, "cm"),
  axis.title = element_blank(),
  legend.background = element_rect(fill = "transparent", color = "transparent"),
  legend.key.height = grid::unit(0.4, "cm"),
  legend.key.width = grid::unit(0.3, "cm"),
  legend.margin = margin(l = 0),
  legend.justification = c("left", "top"),
  legend.spacing.x = grid::unit(0.05, "cm"),
  legend.spacing.y = grid::unit(0.05, "cm"),
  legend.text = element_text(vjust = 0.7, margin = margin()),
  legend.title = element_blank(),
  panel.grid = element_blank(),
  plot.title = element_text(size = rel(1.1), face = "bold")
)

point_colors <- c("#FDE74C", "#EE7733", "#0077BB")

pretty_breaks <- function(x, .axis = "x", dx = 2500, n = 4){

  mn <- paste0(.axis, "min")
  mx <- paste0(.axis, "max")

  brks <- seq(x[[mn]] + dx, x[[mx]] - dx, length.out = n)

  round(brks, digits = -3)

}

```

## Data

### Geopackage

Specify path to geopackage database holding all spatial vector data.

```{r}

gpkg <- here("data", "choyero.gpkg")

```

### Project Window

```{r}

bcs <- read_sf(gpkg, layer = "baja")

window <- read_sf(gpkg, layer = "window")

roads <- read_sf(gpkg, layer = "roads")

watersheds <- read_sf(gpkg, layer = "watersheds")

```

### Elevation (DEM)

```{r}

dem <- rast(here("data", "dem_30m.tif"))

```

### O-D Points

```{r}

# ORIGIN ---
clusters <- 
  read_sf(gpkg, layer = "clusters") |>
  mutate(variable = "cluster")

# DESTINATION --- to city, terminus at edge of window
terminus <- 
  read_sf(gpkg, layer = "terminus") |>
  mutate(variable = "terminus")

springs <- 
  read_sf(gpkg, layer = "springs") |>
  select(id, name) |>
  mutate(variable = "spring")

```

```{r}
#| code-fold: true

od_points <- clusters |> 
  rename("name" = cluster) |> 
  mutate(name = as.character(name)) |> 
  select(name, variable) |> 
  bind_rows(terminus, springs) |> 
  mutate(variable = ifelse(variable == "terminus", "city", variable))

bb8 <- st_bbox(window)
dx <- bb8[["xmax"]] - bb8[["xmin"]]
dy <- bb8[["ymax"]] - bb8[["ymin"]]

# add hillshade relief to visualize elevation
# https://dominicroye.github.io/en/2022/hillshade-effects/
hillshade <- shade(
  slope = terrain(dem, "slope", unit = "radians"),
  aspect = terrain(dem, "aspect", unit = "radians"),
  angle = 45,
  direction = seq(45, 360, by = 45),
  normalize = TRUE
)

hillshade <- mask(sum(hillshade), vect(st_union(watersheds)))
names(hillshade) <- "shade"

gg <- ggplot() +
  geom_raster(
    data = as.data.frame(hillshade, xy = TRUE),
    aes(x, y, fill = shade)
  ) +
  scale_fill_distiller(
    palette = "Greys", 
    na.value = "transparent",
    guide = "none"
  ) +
  new_scale_fill() +
  geom_raster(
    data = as.data.frame(dem, xy = TRUE),
    aes(x, y, fill = elevation),
    alpha = 0.7
  ) +
  scale_fill_distiller(
    palette = "Greys", 
    na.value = "transparent",
    guide="none"
  ) +
  new_scale_fill() +
  geom_sf(
    data = watersheds, 
    fill = "transparent", 
    color = alpha("darkblue", 0.35),
    linewidth = 0.2
  ) +
  geom_sf(
    data = roads, 
    color = "black",
    linewidth = 0.3
  ) +
  geom_sf(
    data = od_points, 
    aes(shape = variable, fill = variable, color = variable),
    stroke = 0.3,
    size = 1.8
  ) +
  scale_color_manual(values = c("grey15", "white", "white")) +
  scale_fill_manual(values = alpha(point_colors, 0.75)) +
  scale_shape_manual(values = c(23, 22, 21)) +
  coord_sf(datum = 4485) +
  scale_x_continuous(
    breaks = pretty_breaks(bb8, "x"), 
    expand = expansion(add = 1000)
  ) +
  scale_y_continuous(
    breaks = pretty_breaks(bb8, "y", n = 3), 
    expand = expansion(add = 1000)
  ) +
  theme(legend.position = c(0.03, 0.98))

ggsave(
  plot = gg,
  filename = here("figures", "od-points.png"),
  dpi = 600,
  width = 3.5,
  height = 3.5 * (dy/dx)
)

```

:::{.mx-auto .my-4 .img-45}
![](../figures/od-points.png)
:::

## Path analysis

To estimate least-cost paths, we first derive slope estimates from the DEM then calculate the hiking speed at which individuals can traverse any given cell using Campbell's hiking function. After that, we decrease the cost in units of time to traverse grid cells crossed by roads and other paths using weights that approximate travel times by car or horse. This is equivalent to increasing the speed one travels in those cells. Below are the specific weights we use.

:::{.w-50 .mx-auto .py-4}
| Type      | Weight | Approx. Speed     |
|:----------|--------|:------------------|
| primary   | 0.2    | 25 mph (41.5 kmh) | 
| secondary | 0.5    | 10 mph (16.2 kmh) |
| tertiary  | 0.8    | 07 mph (11.0 kmh) |

: Down-weight for travel-cost
:::

```{r}

primary_roads   <- roads |> filter(level == "P") |> st_buffer(dist = 30)
secondary_roads <- roads |> filter(level == "S") |> st_buffer(dist = 30)
tertiary_roads  <- roads |> filter(level == "T") |> st_buffer(dist = 30)

terrain <- dem |> 
  hf_terrain(max_slope = 35, decile = 30) |> 
  hf_channel(primary_roads,   .m = 0.2) |> 
  hf_channel(secondary_roads, .m = 0.5) |>
  hf_channel(tertiary_roads,  .m = 0.8)

remove(primary_roads, secondary_roads, tertiary_roads)

```

```{r}
#| code-fold: true

gg <- ggplot() +
  geom_raster(
    data = as.data.frame(hf_rasterize(terrain), xy = TRUE),
    aes(x, y, fill = lyr.1)
  ) +
  scale_fill_viridis(
    name = "Seconds",
    breaks = c(10, 30, 50)
  ) +
  coord_sf(datum = 4485) +
  scale_x_continuous(
    breaks = pretty_breaks(bb8, "x"), 
    expand = expansion(add = 1000)
  ) +
  scale_y_continuous(
    breaks = pretty_breaks(bb8, "y"), 
    expand = expansion(add = 1000)
  ) +
  guides(
    fill = guide_colorbar(
      barwidth = 3.8, 
      barheight = 0.6,
      raster = FALSE,
      frame.linewidth = 0.2,
      ticks.linewidth = 0.7,
      title.position = "top"
    )
  ) +
  theme(
    legend.position = c(0.03, 0.98),
    legend.direction = "horizontal",
    legend.spacing.x = grid::unit(0.05, "cm"),
    legend.text = element_text(size = rel(0.5), margin = margin(l = 2)),
    legend.title = element_text(
      size = rel(0.8), 
      margin = margin(b = 2, t = 2),
      hjust = 0
    )
  )

ggsave(
  plot = gg,
  filename = here("figures", "cost-surface.png"),
  dpi = 600,
  width = 3.5,
  height = 3.5 * (dy/dx)
)

```

:::{.mx-auto .my-4 .img-45}
![](../figures/cost-surface.png)
:::

### To springs

```{r}

to_springs <- hf_hike(
  terrain, 
  from = clusters, 
  to = springs,
  add_cost = TRUE
)

to_springs <-
  to_springs |>
  group_by(from) |> 
  slice_min(cost, n = 2) |> # for each cluster, get two lowest cost paths
  mutate(cost = (cost/3600)) |> # convert to hours 
  summarize(
    cluster = unique(from),
    spring = paste(to, collapse = ","),
    cost = mean(cost)
  ) |> 
  select(cluster, spring, cost)

```

### To cities

Please note that travel time is not strictly to cities, but to the edge of the project area. As noted above, for the northern watershed, the time to La Paz equals the time it takes to leave the watershed plus half an hour.

```{r}

north <- clusters |> 
  st_filter(
     st_union(filter(watersheds, id %in% c(3294, 3296, 3299)))
  )

south <- clusters |> 
  st_filter(
    filter(watersheds, id == 3326)
  )

### NORTH ###
out_north <- hf_hike(
    terrain, 
    from = north, 
    to = terminus[1, ], 
    add_cost = TRUE
  ) |> 
  mutate(
    cluster = north$cluster,
    city = "Ciudad Constitucion", 
    cost = (cost/3600),
    watershed = "north"
  )

### SOUTH ###
out_south <- hf_hike(
    terrain, 
    from = south, 
    to = terminus[2, ], 
    add_cost = TRUE
  ) |> 
  mutate(
    cluster = south$cluster,
    city = "La Paz",
    cost = (cost/3600),
    watershed = "south"
  )

to_cities <- 
  bind_rows(out_north, out_south) |> 
  mutate(time_to_paz = ifelse(watershed == "north", cost + 0.5, cost)) |>
  select(cluster, city, cost, time_to_paz)

remove(north, south, out_north, out_south)

```

### LCP Map

```{r}
#| code-fold: true

paths <- bind_rows(
  to_springs |> mutate(destination = "To springs"),
  to_cities |> mutate(destination = "To cities")
)

gg <- ggplot() +
  geom_raster(
    data = as.data.frame(hillshade, xy = TRUE),
    aes(x, y, fill = shade)
  ) +
  scale_fill_distiller(
    palette = "Greys", 
    na.value = "transparent",
    guide = "none"
  ) +
  new_scale_fill() +
  geom_raster(
    data = as.data.frame(dem, xy = TRUE),
    aes(x, y, fill = elevation),
    alpha = 0.7
  ) +
  scale_fill_distiller(
    palette = "Greys", 
    na.value = "transparent",
    guide="none"
  ) +
  new_scale_fill() +
  geom_sf(
    data = watersheds, 
    fill = "transparent", 
    color = alpha("darkblue", 0.35),
    linewidth = 0.2
  ) +
  geom_sf(
    data = roads, 
    color = "black",
    linewidth = 0.3
  ) +
  geom_sf(
    data = paths |> st_buffer(125) |> group_by(destination) |> summarize(),
    fill = "#C7FFED",
    color = "#00291D",
    linewidth = 0.2
  ) +
  facet_wrap(vars(destination)) +
  geom_sf(
    data = od_points, 
    aes(shape = variable, fill = variable, color = variable),
    stroke = 0.3,
    size = ifelse(od_points$variable == "city", 2.2, 1.5) |> rep(2)
  ) +
  scale_color_manual(values = c("grey15", "white", "white")) +
  scale_fill_manual(values = alpha(c(point_colors, 0.75))) +
  scale_shape_manual(values = c(23, 22, 21)) +
  coord_sf(datum = 4485) +
  scale_x_continuous(
    breaks = pretty_breaks(bb8, "x"), 
    expand = expansion(add = 1000)
  ) +
  scale_y_continuous(
    breaks = pretty_breaks(bb8, "y", n = 3), 
    expand = expansion(add = 1000)
  ) +
  theme(
    legend.position = c(0.01, 0.98),
    strip.background = element_blank(),
    strip.text = element_text(size = rel(1.1), hjust = 0)
  )

ggsave(
  plot = gg,
  filename = here("figures", "least-cost-paths.png"),
  dpi = 600, 
  width = 5.75,
  height = 5.75 * (dy/(2*dx)) + 0.47
)

```

:::{.mx-auto .my-4 .img-65}
![](../figures/least-cost-paths.png)
:::

### Save

```{r}

write_sf(
  to_springs,
  dsn = gpkg,
  layer = "path_to_springs"
)

write_sf(
  to_cities,
  dsn = gpkg,
  layer = "path_to_cities"
)

```

## References

::: {#refs}
:::

## Session Info

```{r}
#| code-fold: true

# save the session info as an object
pkg_sesh <- sessioninfo::session_info(pkgs = "attached")

# inject the quarto info
pkg_sesh$platform$quarto <- paste(
  quarto::quarto_version(), 
  "@", 
  quarto::quarto_path()
  )

# print it out
pkg_sesh

```

### 2023-07-12 R/models.qmd
🚀 initial commit

---
title: "Supplement B: Fitting Models"
date: today
format:
  html: 
    embed-resources: true
    fig-align: center
    fig-dpi: 300
    fig-responsive: true
    number-sections: true
    number-depth: 3
    toc: true
    toc-depth: 3
    smooth-scroll: true
    theme: cosmo
    highlight-style: a11y
execute: 
  echo: true
  warning: false
  error: false
---

<!-- 
RStudio/Quarto is having some issues with figure sizing, 
so using this hacky CSS solution. 
-->

```{css}
#| echo: false

.img-45 {
  width: 4.5in;
}

.img-65 {
  width: 6.5in;
}

```

## Overview

__Goal__: model the spatial distribution of ranch clusters and populations.

__Data__: Dr. Macfarlan's ethnographic data, along with ecological and socio-economic variables.

__Method__: binomial GLM for occupied vs abandoned, negative binomial for population

We want to know what motivates market-integrated, subsistence ranchers to live where they do. This is framed as a trade-off between market integration and ecology.

## R Preamble

```{r libraries}

library(MASS)
library(broom)
library(car)
library(dplyr)
library(ggplot2)
library(here)
library(gt)
library(patchwork)
library(purrr)
library(sf)
library(sfdep)
library(tidyr)
library(viridis)

```

Specify consistent plot theme here:

```{r}
#| code-fold: true

theme_set(theme_bw(12))

theme_update(
  panel.grid = element_blank(),
  plot.margin = margin(2,2,2,2),
  strip.background = element_blank(),
  strip.text = element_text(hjust = 0, size = rel(1.1))
)

```

And some defaults for table formatting:

```{r}
#| code-fold: true

rename_models <- function(x){

  x |> 
    mutate(
      model = case_when(
        model == "occupied"   ~ "Occupied-Abandoned",
        model == "population" ~ "Population",
        model == "secure"     ~ "Population [+ secure]",
        model == "securex"    ~ "Population [x secure]",
        TRUE ~ NA
      )
    )

}

format_double <- function(x){

  x |> mutate(across(where(is.double), \(.x){ round(.x, digits = 3) }))

}

model_table <- function(x){

  x |> 
    gt() |> 
    tab_options(
      column_labels.font.weight = "bold",
      data_row.padding = px(6),
      heading.align = "left",
      heading.border.bottom.style = "hidden",
      row.striping.background_color = "gray96",
      row.striping.include_stub = FALSE,
      row.striping.include_table_body = TRUE,
      table.align = "left",
      table.border.top.style = "hidden",
      table.border.bottom.style = "hidden"
    ) |> 
    sub_missing(missing_text = "")

}

```

## Data

### Geopackage

Specify path to geopackage database holding all spatial vector data.

```{r}

gpkg <- here("data", "choyero.gpkg")

```

### Dependent variables

* Occupied vs Abandoned ranches
* Population size of occupied ranches

Rancheros live in "ranch clusters" occupied by close kin.

```{r}

ranches <- read_sf(gpkg, layer = "ranches")

clusters <- read_sf(gpkg, layer = "clusters")

```

:::{.mx-auto .my-4 .img-65}
![](../figures/overview-map.png)
:::

### Independent variables

* Market-integration: cost-distance (sort of) to La Paz and Constitucion (hours)  
* Ecology: cost-distance to springs (hours)

Cost-distance to La Paz/Ciudad Constitucion provides a proxy measure of access to markets where rancheros sell livestock and buy food and equipment. Cost-distance to springs represents habitat suitability for small-scale pastoralist economies. Measuring it is a complicated matter, as property rights are involved and multiple springs are used. To avoid some of that complexity, we take the average travel time from a ranch cluster to its two nearest springs. So, this is our basic trade-off between market-integration and local ecology. For further details about how we estimate this variable, see "R/leastcostpaths.html."

```{r}

path_to_cities <- read_sf(gpkg, layer = "path_to_cities")

path_to_springs <- read_sf(gpkg, layer = "path_to_springs")

security <- ranches |>   
  st_drop_geometry() |>
  select(cluster, land) |> 
  mutate(secure = ifelse(land == "secure", 1, 0)) |> 
  group_by(cluster) |> 
  summarize(secure = sum(secure)/n())

clusters <- 
  clusters |> 
  mutate(
    springs = path_to_springs$cost,
    cities = path_to_cities$cost,
    paz = path_to_cities$time_to_paz,
    secure = security$secure
  )

remove(path_to_cities, path_to_springs, security, ranches)

```

## Analysis

### Models

In addition to a binomial model of the status of each ranch cluster (occupied vs abandoned), we fit a negative binomial model of the population count of occupied ranch clusters. The negative binomial is an alternative to standard Poisson models where there is an expectation of dispersion in the residuals. The Poisson model assumes that the variance is equal to the mean, but this is not always the case.

```{r}

models <- list(

  occupied = glm(
    occupied ~ paz + springs,
    family = binomial,
    data = clusters
  ),
  population = glm.nb(
    population ~ cities + springs,
    data = clusters |> filter(population > 0) 
  ),
  secure = glm.nb(
    population ~ (cities + springs) + secure,
    data = clusters |> filter(population > 0) 
  ),
  securex = glm.nb(
    population ~ (cities + springs) * secure,
    data = clusters |> filter(population > 0)
  )

)

```

### ANOVA (LRT)

We conduct a Likelihood Ratio Test of the population models to see if adding land security makes a significant improvement to the model.

```{r}

lrt <- with(models, anova(population, secure, securex, test = "LRT"))

```

### Residual Autocorrelation

We use Monte Carlo simulations of Moran's I to test for spatial autocorrelation in the untransformed residuals of each model.

```{r}

get_moran <- function(x, .sf){

  is_nb <- grepl("Negative Binomial", x$family[["family"]])

  if (is_nb) .sf <- .sf |> filter(occupied == 1)

  geometry <- st_geometry(.sf)

  neighbors <- st_knn(geometry, k = nrow(.sf)-1)
  weights <- st_inverse_distance(neighbors, geometry)

  moran <- global_moran_perm(residuals(x), neighbors, weights)

  moran |> 
    tidy() |> 
    mutate(variable = "residuals", .before = 1)

}

morans_i <- models |> 
  map(\(.x){ get_moran(.x, clusters) }) |> 
  bind_rows(.id = "model")

```

### Variance Inflation

A model of variance inflation tests for multi-collinearity in the predictors.

```{r}

vif <- models |> 
  map(vif) |> 
  bind_rows(.id = "model")

```

## Results

### Model evaluation

```{r}
#| code-fold: true

model_statistics <- models |> 
  map(glance) |> 
  # glance returns logLik column as numeric for glm and 'logLik" for negbin
  map(\(.x){ mutate(.x, across(everything(), as.numeric)) }) |> 
  bind_rows(.id = "model") |> 
  format_double() |> 
  rename_models() |> 
  rename("n.obs" = nobs) |> 
  relocate(model)

model_statistics |> 
  model_table() |> 
  tab_header(title = "Model Statistics")

```

```{r}
#| code-fold: true

coefficient_estimates <- models |> 
  map(tidy) |> 
  bind_rows(.id = "model") |> 
  format_double() |> 
  rename_models() |> 
  rename("z value" = statistic) |>
  rename_with(gsub, pattern = "\\.", replacement = " ")

coefficient_estimates |> 
  group_by(model) |> 
  model_table() |> 
  tab_options(row_group.as_column = TRUE) |> 
  tab_stubhead("model") |> 
  tab_header(title = "Coefficient Estimates")

```

```{r}
#| code-fold: true

lrt <- lrt |> 
  as_tibble() |> 
  rename_with(tolower) |> 
  format_double() |> 
  mutate(
    model = case_when(
      model == "cities + springs"            ~ "Population",
      model == "(cities + springs) + secure" ~ "Population [+ secure]",
      model == "(cities + springs) * secure" ~ "Population [x secure]",
      TRUE ~ NA
    )
  )

lrt |> 
  model_table() |> 
  tab_header(
    title = "Likelihood Ratio Test",
    subtitle = "Alternative: difference in log-likelihood greater than zero"
  )

```

```{r}
#| code-fold: true

morans_i <- morans_i |> 
  format_double() |> 
  rename_models() |> 
  rename("rank" = parameter)

morans_i |> 
  select(model, statistic, rank, p.value) |> 
  model_table() |> 
  tab_header(
    title = "Monte Carlo Simulations of Moran's I",
    subtitle = "Alternative: spatial autocorrelation (two-sided)"
  )

```

```{r}
#| code-fold: true

vif <- vif |> 
  format_double() |> 
  rename_models()

vif |> 
  model_table() |> 
  tab_header(
    title = "Variance Inflation",
    subtitle = "Alternative: "
  )

```

High VIF scores for interaction terms can be ignored.

```{r}
#| echo: false

save(
  model_statistics, coefficient_estimates, morans_i, vif,
  file = here("data", "model-statistics.rdata")
)

```

### Marginal Response

Here we calculate the marginal response of each response variable relative to change in each focal variable while holding all non-focal variables at their mean.

```{r}

# to make the lines smoother when plotting
densify <- function(x, n = 300) seq(min(x), max(x), length = n)

template <- with(
  clusters, 
  tibble(
    paz = densify(paz),
    cities = densify(cities),
    springs = densify(springs)
  )
)

estimate <- function(model, newdata){

  est <- predict(
    model, 
    newdata = newdata, 
    se.fit = TRUE
  )

  inv_link <- family(model)$linkinv

  est <- with(
    est,
    tibble(
      y  = inv_link(fit),
      hi = inv_link(fit + 2*se.fit),
      lo = inv_link(fit - 2*se.fit)
    )
  )

  bind_cols(newdata, est)

}

responses <- tibble(
    model = c("occupied", "occupied", "population", "population"),
    variable = c("paz", "springs", "cities", "springs")
  ) |> 
  mutate(
    model_object = models[model],
    data = list(template),
    data = map2(data, variable, \(.x, .y){ .x |> mutate(across(-all_of(.y), mean)) }),
    estimate = map2(model_object, data, estimate),
    estimate = map2(estimate, variable, \(.x, .y){ rename(.x, "x" = all_of(.y)) })
  ) |> 
  select(model, variable, estimate) |> 
  unnest(estimate) |> 
  select(model, variable, x, y, hi, lo)

```

```{r}
#| code-fold: true
#| fig-asp: 1.1

# add old data to display points
observations <- map(models, \(.x){ .x$model })

observations <- tibble(
    model = c("occupied", "occupied", "population", "population"),
    variable = c("paz", "springs", "cities", "springs")
  ) |> 
  mutate(
    old_data = observations[model],
    old_data = map2(old_data, variable, \(.x, .y){ rename(.x, "x" = all_of(.y)) }),
    old_data = map2(old_data, model,    \(.x, .y){ rename(.x, "y" = all_of(.y)) })
  ) |> 
  unnest(old_data) |> 
  select(model, variable, x, y)

baja_labels <- c(
    "paz" = "To La Paz",
    "springs" = "To Springs",
    "cities" = "To Cities"
)

gg_occupied <- ggplot() +
  geom_ribbon(
    data = responses |> filter(model == "occupied"),
    aes(x, ymin = lo, ymax = hi),
    fill = "#F5F5F5",
    color = "transparent"
  ) +
  geom_line(
    data = responses |> filter(model == "occupied"),
    aes(x, y),
    linewidth = 0.7
  ) +
  geom_point(
    data = observations |> filter(model == "occupied"),
    aes(x, y),
    size = 2.5,
    color = alpha("#2E4756", 0.7)
  ) +
  labs(
    x = "Hours",
    y = "Abandoned-Occupied"
  ) +
  facet_wrap(
    vars(variable), 
    scales = "free_x",
    labeller = labeller(variable = baja_labels)
  ) +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.title.x = element_blank()
  )

gg_population <- ggplot() +
  geom_ribbon(
    data = responses |> filter(model == "population"),
    aes(x, ymin = lo, ymax = hi),
    fill = "#F5F5F5",
    color = "transparent"
  ) +
  geom_line(
    data = responses |> filter(model == "population"),
    aes(x, y),
    linewidth = 0.7
  ) +
  geom_point(
    data = observations |> filter(model == "population"),
    aes(x, y),
    size = 2.5,
    color = alpha("#2E4756", 0.7)
  ) +
  labs(
    x = "Hours",
    y = "Population"
  ) +
  facet_wrap(
    vars(variable),
    scales = "free_x",
    labeller = labeller(variable = baja_labels)
  )

gg <- gg_occupied/gg_population

ggsave(
  plot = gg,
  here("figures", "response-plots.png"),
  dpi = 600,
  width = 5.75,
  height = 5.75
)

```

:::{.mx-auto .my-4 .img-45}
![](../figures/response-plots.png)
:::

## Session Info

```{r}
#| code-fold: true

# save the session info as an object
pkg_sesh <- sessioninfo::session_info(pkgs = "attached")

# inject the quarto info
pkg_sesh$platform$quarto <- paste(
  quarto::quarto_version(), 
  "@", 
  quarto::quarto_path()
  )

# print it out
pkg_sesh

```

### 2023-07-12 manuscript/manuscript.qmd
🚀 initial commit

---
title: Market integration mediates human-environment interactions among traditional ranchers in Baja California Sur, Mexico
date: today
author:
  - name: Kenneth B. Vernon
    affiliations: 
      - id: colorado
        address: | 
          Center for Collaborative Synthesis in Archaeology  
          Institute of Behavioral Science  
          University of Colorado, Boulder  
          1440 15th Street  
          Boulder, CO 80302 
  - name: Brian F. Codding
    affiliations:
      - id: utah-anthropology
        address: | 
          University of Utah Department of Anthropology, 
          260 S. Central Campus Drive, RM 4625, 
          Salt Lake City, UT 84112
      - id: gcsc
        address: |
          University of Utah Global Change and Sustainability Center, 
          115 South 1460 East, RM 234 FASB, 
          Salt Lake City, UT 84112
  - name: Simon Brewer
    affiliations:
      - id: utah-geography
        address: |
          University of Utah Department of Geography, 
          260 S. Central Campus Drive, RM 4625, 
          Salt Lake City, UT 84112
  - name: Shane J. Macfarlan
    affiliations:
      - ref: utah-anthropology
      - ref: gcsc
target-journal:
  - "Nature Sustainability"
  - "Human Ecology"
  - "PNAS"
abstract: |
  Market integration presents new opportunities and challenges for traditional communities still engaged in a largely subsistence-based economy. In this paper, we explore one particular aspect of that dynamic: how market integration affects the way that populations of historically marginalized ranchers distribute themselves across a landscape, where they feel the competing pull of urban amenities and local ecosystem services. Our study focuses on a small community of ranchers living in kin-based ranch clusters in the rugged and arid landscapes of the Sierra de la Giganta mountains along the eastern spine of the Baja California peninsula in Mexico. Our models show that ranch clusters are more likely to be abandoned the farther they are from surficial springs, and occupied ranch clusters have larger populations the closer they are to markets. This has important consequences for the ecological resilience of this ranching community, particularly in the context of climate change.
keywords:
  - Market Integration
  - Ideal Free Distribution
  - Behavioral Ecology
  - Sustainability
  - Urbanization
  - Land Abandonment
corresponding_author: Kenneth.Vernon@colorado.edu
linenumbers: false
watermark: false
bibliography: bibliography.bib
format:
  simple-pdf: default
execute: 
  echo: false
  error: false
  warning: false
---

```{r}

library(dplyr)
library(gt)
library(here)
library(knitr)

```

# Introduction

In traditional, small-scale subsistence economies, the value of a habitat is typically a function of its local resources, particularly their distribution and abundance. As individuals integrate with modern markets, however, that connection becomes more tenuous. This can be a good thing, of course, allowing individuals to diversify their livelihood, making them more flexible and capable of absorbing potential risks and uncertainty [@barrett2001; @little2001]. But, it can have other, unforeseen consequences, too, especially as individuals come to depend on markets for their livelihood. For those living in predominantly marginal habitats, it may even have the opposite effect, making them more sensitive to the effects of climate change and other exogenous forces like infrastructure development and range fragmentation [@fratkin2001; @hobbs2008; @thornton2009]. Together, these conflicting outcomes suggest important trade-offs for individuals presented with the opportunity to integrate more closely with modern markets. Time and energy they put into accessing regional market services is often time and energy they cannot spend accessing local ecosystem services. How they navigate that trade-off can have important consequences for their health and well-being [see, for example, @liebert2013; @urlacher2016; @stagaman2018], including dramatic changes to their life history and demography [@shenk2016; @colleran2015]. It is, thus, critically important to understand how and to what degree individuals choose to integrate with modern markets.

Here we explore that question, focusing on a small group of historically marginalized rural ranchers living in the arid Sierra de la Giganta region of Baja California Sur, Mexico [@macfarlan2020a, see @fig-overview-map]. While these people are sometimes referred to as "choyeros", a local nickname for anyone born in the rural, mountainous ranching communities of Baja California Sur (also a reference to the local cholla cactus, *Cylindropuntia cholla*), in this paper, we refer to them simply as "ranchers" or "rancheros", which are the expressions they typically use to describe themselves. Descended from Euro-American colonists who began settling the region in the late-17th century [@crosby1994], they number approximately 4,000 today, living in communities largely organized around clusters of one or more ranches housing inter-generational kin. These communities are some of the most severely under-served in the region, lacking access to critical services and infrastructure, including piped water, sanitation, and energy [@secretaríadedesarrollosocial2021].

Given the aridity of the area and lack of regionally connective water infrastructure, rancheros depend almost entirely on surficial springs to satisfy their water needs, including drinking water for themselves and their livestock, as well as irrigation for garden plots (known as huertas and jardines), which provide a small supplement to human and livestock diets during dry periods [@macfarlan2020a; @macfarlan2020b; @schniter2021; @lerback2022]. To make use of springs, ranchers either have to pump water themselves, using gas or solar powered equipment, or transport it by gravity irrigation [@lerback2022]. A strong incentive thus exists to locate in closer proximity to these water resources.

A similar, though competing, incentive exists for accessing markets in urban areas like Loreto, Ciudad Constitución, and La Paz, the capital of Baja California Sur. These cities furnish rural ranchers with many of the necessities they cannot acquire locally, including the ability to trade livestock, register animals with the state, receive medical attention, renew water permits, purchase groceries, and socialize with people outside their local community. The chance for supplemental income will also on occasion draw rancheros to the cities, as it has for pastoralists elsewhere [@barrett2001; @little2001]. This includes working in construction, seasonal employment in coastal fisheries, or the tourist sector, specifically within the resort towns of Cabo San Lucas and San Jose del Cabo.

As a baseline for investigating the demographic consequences of these competing necessities, we first explore how market integration and local ecology structure the distribution of occupied and abandoned ranch clusters. This is a process that took place over several hundred years, with the shift to its current distribution beginning in the late 1970s, following construction of the transpeninsular Federal Highway 1 [@barkenbus1974]. We then ask how market integration and local ecology have shaped the distribution of the population across occupied clusters today.

```{r}
#| label: fig-overview-map
#| fig-align: center
#| out-width: 5.75in
#| fig-cap: "**Overview map.** The top right panel shows the location of the study area (in red) in relation to the two nearest city centers, La Paz and Ciudad Constitucion. Panel A shows the distribution of ranch clusters in each watershed, as well as their population size (size and color). Note that a population of zero means the ranch cluster is abandoned. Panel B shows the distribution of land secure and land insecure ranch houses (shape and color) for all occupied ranch houses. Solid white lines represent Federal Highway 1 (in the top right panel), major roads connecting it to the project area, and other roads and paths in the project area."

here("figures", "overview-map.png") |> include_graphics()
```

Importantly, we investigate the degree to which modern land fragmentation, specifically the security of land title, serves as an effective constraint on these dynamics. This is important because pastoralists have historically lacked clear title to land, instead relying on open rangeland systems [@behnke1994; @turner1999]. In Mexico, these systems take the form of *ejidos*, or areas of communal agricultural land. The first *ejidos* were established by the Mexican state in the early part of the twentieth century [@perramond2008], though in our project area they did not arrive until 1975 [@dof1975] and 1976 [@dof1976], respectively. While these *ejidos* have historically served as a buffer against land fragmentation, allowing rancheros to maintain their way of life, their ability to play that role continues to change as the pace of privatization and shifting land tenure accelerates [@lesorogol2003; @reid2008]. As has happened in other contexts, privatization efforts are often led by NGOs, commercial interests, and other external forces, with the result being greater fragmentation of pastoral rangeland systems [@burnsilver2008; @galvin2009]. For the rural ranchers of the Sierra de la Giganta, this process of fragmentation at least partially explains why ranches of related individuals tend to cluster. Rancheros, unable to find new land for themselves, simply build ranches next to their kin on land that is at least historically associated with their family.

Our analysis relies on two generalized linear models (GLMs). We use a binomial GLM to evaluate whether market integration or local ecology has played a larger role in determining the current status of ranch clusters, if they are occupied or abandoned. We then use a negative binomial GLM to identify which variable contributes more to the distribution of these ranching populations across currently occupied clusters, with a binary term representing land security. These are applied to ranch clusters in two east-west running watersheds along the southern extent of the Sierra de la Giganta range.

The key measure in all these models is cost-distance or travel time. This has been used as a proxy for market integration in other studies focusing on largely subsistence-based economies [see especially @gurven2017], with models exploring its relationship to, among other things, social status [@schultz2019], hunting effort and efficiency [@luz2015], and oral tradition [@schniter2018]. An important advantage of our analysis, however, is that we use cost-distance as a proxy for both market integration and local ecology (or ecosystem services). Specifically, we use cost-distance to major cities as a proxy for market integration and cost-distance to local surficial springs as a proxy for ecosystem services. This allows for more straightforward comparisons of their relative contributions to settlement outcomes, which can then be readily interpreted as strategies for navigating trade-offs between those two variables.

We would also like to emphasize an important historical dimension to this research, namely, that the process leading to the current distribution of occupied and abandoned ranch clusters played out before Ciudad Constitucion (the large urban area to the north of our study area) offered significant market benefits. This is reflected in how we specify each model. The model of occupied and abandoned ranch clusters includes only proximity to La Paz as a measure of market integration, which historically was the only market that ranchers accessed prior to the construction of Federal Highway 1. The model of population distributions across occupied ranches, however, includes cost-distance to La Paz and Ciudad Constitucion, as both have offered markets regularly used by ranchers in our study population today. All models are evaluated for goodness of fit, multicollinearity, and residual autocorrelation.

# Results

We initiated this research to understand how local ecology and market integration influenced settlement dynamics. Our first model examines how these two factors influence whether or not a ranch cluster is occupied or abandoned. The model suggests that a cluster's status (whether it is occupied or abandoned) negatively co-varies with cost-distance to spring ($\beta$ = -4.803, *p* = 0.009) and positively co-varies with cost-distance to La Paz ($\beta$ = 2.797, *p* \< 0.001), indicating that ranch clusters are more likely to be occupied closer to springs and further from town (@tbl-models, @fig-response-plots). Our second model examines how local ecology and market integration influence the distribution of the population across occupied ranch clusters today. The model shows that population size across occupied ranch clusters negatively co-varies with cost-distance to cities ($\beta$ = -0.802, *p* \< 0.001), for both Constitucion and La Paz, but does not significantly co-vary with cost-distance to springs ($\beta$ = 0.699, *p* = 0.299) (@tbl-models, @fig-response-plots). The model with an additive land security term suggests that more secure ranch clusters have slightly smaller populations, but the effect is non-significant ($\beta$ = -0.114, *p* = 0.687). The ANOVA also indicates that an additive land security term does not significantly improve the model ($\chi$(1) = 3.003, *p* = 0.684). The same is true for an interactive term.

The null hypothesis of no spatial autocorrelation in the untransformed residuals cannot be rejected for either the occupation status (*I* = -0.023, *p* = 0.820) or population size (*I* = -0.082, *p* = 0.460) model, thus indicating a lack of spatial correlation. Measures of variance inflation are minimal in each case, approximately one for all terms, strongly suggesting no multicollinearity in the independent variables.

```{r}
#| label: tbl-models

here("data", "model-statistics.rdata") |> load()

coefficient_estimates |>
  filter(model %in% c("Occupied-Abandoned", "Population")) |>
  group_by(model) |> 
  gt() |> 
  tab_options(
    column_labels.font.weight = "bold",
    data_row.padding = px(6),
    heading.align = "left",
    heading.border.bottom.style = "hidden",
    row_group.as_column = TRUE,
    row.striping.background_color = "gray96",
    row.striping.include_stub = FALSE,
    row.striping.include_table_body = TRUE,
    table.border.top.style = "hidden",
    table.border.bottom.style = "hidden"
  ) |> 
  sub_missing(missing_text = "") |> 
  tab_stubhead("model") |> 
  tab_header(title = "Coefficient Estimates") |> 
  tab_caption("") |> 
  as_latex()

```

# Discussion

With this analysis, we seek to understand how rural ranchers in Baja California Sur respond to important trade-offs between their local ecology and market integration. Our results indicate that the locations of modern ranch clusters are largely constrained by access to surficial water sources, our proxy for habitat suitability in this arid environment, rather than access to the capital La Paz. This observation is supported by the fact that occupied ranch clusters are more likely to be found farther from the capital, rather than closer to it. However, within those constraints, the rancheros are now living at larger densities in locations with easier access to Ciudad Constitucion and La Paz, our proxies for market integration. In other words, water availability determines *where* people live but not *how many* people live there. Instead, that is being driven largely by markets, which pull people closer to the cities.

```{r}
#| label: fig-response-plots
#| fig-align: center
#| out-width: 5in
#| fig-cap: "**Response plots.** Marginal response plots showing change in the dependent variables (occupied vs abandoned ranch clusters and population size) as a function of change in one independent variable while holding the other at its mean."

here("figures", "response-plots.png") |> include_graphics()
```

## Proxies for market integration

One important advantage of this research concerns the use of travel costs as a measure of both market integration and local adaptation. Using this simple metric provides a straightforward way of evaluating the relative contribution of each variable to the outcome of interest, in this case, settlement decisions among rural ranchers. It also conforms with geographic theory regarding spatial interactions and dependence (otherwise known as Tobler's Law) and overcomes known limitations of Euclidean distance as a cost measure. And, it is a simple matter to derive those cost estimates from easily accessible remote-sensed elevation and road network data held in online repositories. High resolution informant data, on the other hand, are generally difficult and costly to acquire, often with little chance of providing the right sort of overlap between different variables required for fair comparisons.

It is true, as @mattison2022 point out, that market integration is a multi-dimensional process, making it at first blush an awkward concept to try and capture in a single variable like cost-distance or caloric-intake [@henrich2010]. Still, Tobler's Law would seem to suggest that all the variables strongly associated with market integration will also co-vary with proximity to cities and their markets. In other words, the costs and benefits of market integration will positively co-vary with access to markets - both will increase as one gains greater access to and interacts more often with markets. So, the closer individuals are, the more likely they should be to derive their caloric requirements from cities, to work in cities, to own televisions, cellphones, and other goods sold in cities, but also the more likely they should be to face intense competition for jobs, higher rents, and other costs associated with living in that proximity.

The ranchers in our study population also recognize that those who live closer to the trans-peninsular highway have greater access to markets and are more likely to interact with them. They have said as much during our conversations with them. It is perhaps worth noting in this context that cost distance (specifically transport cost) is also a fundamental variable in the new economic geography, where models seek to explain urbanization or the spatial concentration of economic activity [@krugman1991; @fujita2001] and thus finds support from that direction, too. So, when we look at where individuals actually choose to live (how near or how far from cities and markets they choose to reside), it is safe to assume we are also getting an answer to the question of how they weigh the costs and benefits of cities against those afforded by a rural way of life.

## Historical antecedents

What explains our results? There are many dimensions to this question, but key to any general understanding of such processes is the idea that people will choose to live where they can make the best living for themselves given the socioeconomic and environmental constraints that confront them. Those constraints will, in turn, differ depending on how one chooses to go about making a living [@vernon2022]. For the people of the Sierra de la Giganta, that means living where you can expect to do ranching and subsistence crop production reasonably well, that is, where one has access to permanent water sources, as well as flat lands that sit above dry riverbeds (so they are not impacted by flooding caused by seasonal tropical storms and hurricanes). Much like the vast majority of the Baja California peninsula, the Sierra de La Giganta is a hot, arid, mountainous landscape with minimal surficial freshwater stocks [@maya1997]. The only permanent sources of naturally occurring freshwater in this region are arid-land springs [@grismer1993; @macfarlan2020a; @macfarlan2020b; @macfarlan2021]. Due to the lack of infrastructure in this region, households require springs to support both domestic (e.g. bathing and drinking) and economic needs (e.g. supporting livestock and crop production). Our statistical results are consistent with our ethnographic interviews with ranchers which indicate that access to a spring is the single most important factor for establishing and maintaining a home [@lerback2022].

Our first statistical model indicates that abandoned ranches were located farther from springs. Although this appears to run counter to our previous claim about the importance of springs to all settlement in the region, we believe this is an artifact of our analytic framework, which required us to collapse the temporal dimension of this largely historical process. Historical and ethnographic evidence suggest that as the first Euro-American colonists entered this region in the 18th century, they occupied the largest springs in close proximity to arable land [@crosby1994]. As the population expanded over time, individuals would move up or down drainages, to the next closest spring that could support subsistence needs. After time, ranchers likely saturated the landscape by occupying all areas with permanent freshwater [sensu @crosby1981] as well as less desirable locations which were supported not by springs, but by ephemeral freshwater sources such as tinajas and seasonally filled lake basins, which can hold water for months or years if they receive sufficient precipitation via the North American Monsoon.

However, after the construction of the Transpeninsular Highway and associated graded dirt roads in the 1970s and 1980s, the payoffs structuring settlement decisions were likely altered, as novel constraints and opportunities emerged. For a subsistence population undergoing market transition, the ideal household location has become both next to a spring and close to a road. These roads allowed ranchers to access urban goods and services more easily and created an incentive for households occupying springs and ephemeral freshwater sources far from the road to move closer to it. This process resulted in a depopulation event between the two major drainages. As this process has unfolded over the last 30-40 years, springs near roads with the closest access to urban markets have experienced increasing intensification as suitable habitats for establishing new households have been exhausted, resulting in the highest population densities.

## Socio-ecological consequences

The findings presented here have important consequences for traditional ranching communities living in similar circumstances, for how they maintain their way of life while meeting the challenges posed by climate change, land fragmentation, and other exogenous forces. For instance, researchers have shown that market integration within traditional agricultural communities involves a shift from intensive to extensive kinship systems, or shifts from small kin networks with higher degrees of relatedness to large kin networks with lower degrees of relatedness [@peña2002]. This appears to be the result of a decreasing reliance on consanguineous or close-kin marriage customs, with families favoring expansive social networks over stronger family ties within a market economy [@shenk2016]. Currently, the ranching communities of Baja California Sur are characterized by more intensive kin networks, but the limited available evidence suggests that this might be changing, with individuals increasingly marrying outside of the ranching community, most likely to individuals living within the cities. Families are as a consequence increasing their connections to the cities and by extension increasing the incentives for integrating more closely with their markets.

These settlement dynamics have also had variable ecosystem outcomes. Springs that exist closer to the road, which are more heavily utilized, typically experience substantial negative outcomes relative to springs that exist in the depopulated zone, including biodiversity loss, higher rates of introduced exotic species, soil degradation, and noise and chemical pollution [@leóndelaluz2006]. This is unfortunate, as the springs and the emergent terrestrial ecosystems they support are considered Ramsar sites (wetlands of international importance), as well as national terrestrial and hydrological priority regions [@arriagacabrera2002; @ramsar2022]. However, it is inaccurate to assume, as many conservation biologists, NGOs, and urban elites seem to do, that ranchers and their lifestyles have uniformly negative impacts on regional ecosystem health [@riemann2005]. For it appears that small, riparian ecosystems surrounding springs in the depopulated hinterland away from roads are currently on the rebound, with higher rates of endemism, and this in spite of the fact that ranching continues in the region. Similar outcomes have also been noted elsewhere [@vaccaro2021]. So, concentration of the population within this region might actually be beneficial, at least from the perspective of biodiversity and ecosystem health, a pattern that appears to be recurring across the globe as populations become more urbanized [@daskalova2023]. For that matter, current research suggests that the complete abandonment of a region can have negative consequences for biodiversity given the co-evolutionary histories of small-scale subsistence economies and their local ecosystems [@fischer2012].

Current demographic trends across the region may also exacerbate these local processes [@vasco2017]. Since the signing of NAFTA, the total population of the southern peninsula has more than doubled, from 375 thousand in 1995 to nearly 800 thousand in 2020. As one might expect, these increases have occurred almost exclusively within the peninsula's urban corridors [@inegi2023a; @inegi2023] . This entails increased demand for goods and services and by necessity, increased demand for the labor that can provide them, thereby increasing the incentives for locating in closer proximity to cities and abandoning more remote locations.

## Climate change and drought

Of particular urgency for the rural ranchers of the Sierra de la Giganta is ongoing climate-induced drought [@cavazos2012; @colorado-ruiz2018]. For the reasons we have emphasized throughout, a lack of access to government services and infrastructure makes them uniquely sensitive to these local impacts [@hansen2019]. That is almost certainly why access to water plays a larger role in constraining the location of occupied ranch clusters. It may also explain the decisions ranchers make about the size and composition of their herds, as intra- and inter-annual variation in precipitation - in the absence of water infrastructure - is the primary factor driving the abundance and distribution of most animal fodder in the region [@pettorelli2013; @macfarlan2023]. Previous research has shown that pastoralists respond to and oftentimes prepare for drought conditions by reducing their herd size, selling off stock in the nearby cities, or switching to more drought tolerant alternatives like small ruminants [@mace1989; @mace1990; @mace1993]. While these animals are more efficient in terms of their total land area, feed, and water requirements, there is less demand for them in city markets. This implies that ranchers in the region, while trying to address the ongoing threat of droughts, will lose important sources of capital they would need to access other resources provided by modern markets that they depend on.

## Conclusion

Trade-offs between local ecology and market integration will continue to shape the development of rural, largely subsistence-based communities, especially as globalization and urbanization accelerate. This is as true of the rural ranchers of the Sierra de la Gigantes as it is for ranchers, pastoralists, and small-scale agriculturalists in other parts of the world. For ranchers living in the remote corners of Baja California Sur, the data coupled with informant reporting together suggest that these ranchers feel the pull of markets, being compelled to infill around ranches nearer to roads that give them greater access to those markets. The consequences of this spatial sorting of the population are complicated as well. While integration can help to mitigate local risks, it also raises new ones, with population packing increasing pressure on nearby water resources that are crucial to the sustainability of these communities. At the same time, the concentration of the population appears to offer positive results for the ecology of abandoned areas, among other things increasing their biodiversity.

The research presented in this paper raises a number of important issues that deserve greater attention, particularly the manner and extent to which proximity or cost-distance to markets correlates with other proxies for integration. Answering that question would go along way towards helping researchers overcome many of the methodological challenges associated with richly ethnographic studies of remote and largely subsistence-based economies and in the process offer new avenues for investigating the ongoing integration of those systems with modern markets.

# Methods

To evaluate the determinants of ranch occupation and population size, we rely on two generalized linear models (GLMs). We use a binomial GLM to evaluate whether market integration or local ecology has played a larger role in structuring the distribution of occupied and abandoned ranch clusters. We then use a negative binomial GLM to identify which variable contributes more to the distribution of rancher populations across currently occupied clusters, with an additional variable representing land security, or the proportion of ranch houses within a cluster having clear title to their property, as this may mediate ranch population size. The use of a negative binomial rather than the standard Poisson model for count data is preferred here to account for dispersion, or circumstances in which the variance deviates from the mean, which violates a key assumption of the Poisson distribution. These models are applied to ranch clusters in two largely east-west running watersheds along the southern extent of the Sierra de la Giganta range.

## Data

The dependent variables, occupation status and population, we obtain from Mexico's online Catálogo do Localidades [@secretaríadedesarrollosocial2021], coupled with fieldwork conducted over several years beginning in 2014. In total, the sample includes 72 ranch clusters, with 39 being unoccupied or abandoned and the remaining 33 being occupied by 290 individuals.

The independent variables, market integration and habitat suitability, we operationalize using three estimates of cost-distance: (i) one to La Paz, (ii) one to Ciudad Constitucion or La Paz (depending on the watershed), and (iii) one to local springs, with each involving a mix of pedestrian and automotive travel. The first two variables are our proxies for market integration, under the assumption that increased interaction with markets will positively covary with proximity to the cities. The third variable, cost-distance to springs, is our proxy for habitat suitability, as water is a critical resource in this arid environment. In the binomial GLM, which examines the distribution of abandoned versus active ranching clusters, we use only cost-distance to La Paz (i) as Ciudad Constitucion did not offer a significant market when those locations were being established. We then use cost-distance to both Ciudad Constitucion and La Paz (ii) in the negative binomial GLM examining ranch-cluster population sizes. Cost-distance to springs (iii) is used in both models, on the assumption that the location of active streams would not have changed during the period of time in question [@grismer1993].

Land fragmentation is here operationalized as land security, defined as the proportion of ranch houses within a cluster with clear title to land or who are registered members of an ejido. Information on this is provided by Mexico's National Institute of Statistics and Geography [@registroagrarionacional2019] and by ethnographic observation. Of the occupied ranch clusters, only 8 are fully secure, meaning all ranch houses within the cluster have full title to their property. Of the remaining 25, less than half of their ranch houses are secure.

## Travel time

To estimate travel time, we first derive a gridded cost surface from a 30-meter resolution digital elevation model derived from the National Elevation Dataset provided by the United States Geological Survey. This is achieved by applying Campbell's hiking function [@campbell2019] to slope estimates extracted from the elevation model. The result is a cost surface where the value of each 30-meter grid cell represents the time in seconds it would take to traverse that cell.

Three important assumptions underlie the cost-surface model. First, we assume that the average walking speed of a rancher is approximately equal to the third decile of US hikers (about 2.5 meters per second over flat ground), which increases or decreases based on slope. Second, we assume that no individual will traverse a slope greater than 35 degrees. Obviously, there will be exceptions to this, but ethnographic observation does suggest that it is true in general for the rancheros in this region. Plus, our distance measure is used for relative comparisons, making the precise absolute value of limited concern. Finally, we assume that travel on roads and built paths will be less costly than through grid cells where vegetation and other landform characteristics provide obstacles to travel. In the cost surface, this is achieved by first classifying paths as primary (for maintained roads), secondary (for non-maintained roads), and tertiary (for walking paths), and then weighting the travel cost associated with each by 0.2, 0.5, and 0.8, respectively. This is equivalent to increasing the speed of travel to 12.5, 5, and 3.125 m/s, making the speeds more comparable to driving a car or riding a horse along the primary and secondary paths, which is what the ranchers here do when going to the city or visiting their springs or neighbors. All road paths were digitized manually using Google Earth and validated through field observation.

```{r}
#| label: fig-shortest-paths 
#| fig-align: center 
#| out-width: 5.75in 
#| fig-cap: "**Shortest paths.** Maps showing least-cost paths to springs and cities."

here("figures", "least-cost-paths.png") |> include_graphics()
```

Once we have this cost surface, we then use Djikstra's algorithm to identify least-cost paths from ranch clusters to springs and cities and, thus, the total time it takes to travel from each origin to each destination. These are shown in @fig-shortest-paths. We note, however, that strictly speaking, we do not measure travel time to the cities, just to the edge of the study region in the direction of Federal Highway 1, for both the northern and southern watersheds. Each one has a single road that connects it to Highway 1 and thus to the markets in Ciudad Constitucion to the north and the capital La Paz to the south. The "road" that connects the watersheds to each other is not maintained and thus rarely used. As a consequence, nearly all of the variance in travel time to cities is found within each watershed. For that reason, we estimate the time it takes to leave each watershed ("leaving" here means hitting the edge of the project area), rather than the true time it takes to get to each city. We interpret this as the relative difference in time it takes to get to Ciudad Constitucion from the northern watershed and the relative difference in time it takes to get to La Paz from the southern watershed. To get to La Paz from the northern watershed, we simply add an arbitrary half hour to the total time to get out of that watershed. The same would presumably be the case to get to Ciudad Constitucion from the southern watershed, though that value does not factor into our analysis. For springs, our ethnographic data reveal that individuals residing within the same ranch cluster tend to rely on one or two springs at most [@lerback2022]. Thus, we estimate road distance to the two nearest springs and take the average.

## Model evaluation

For completeness, we fit three versions of the negative binomial model of ranch cluster population: one without the land security variable, one with it included as an additive term, and one with it included as an interaction term. After fitting these models, we perform an ANOVA Likelihood Ratio Test for significant improvement in fit. We then test for spatial autocorrelation in the untransformed residuals of all models, including the binomial model, using Monte Carlo simulations of Moran's I. We also test for multicollinearity in our cost-distance variables by calculating their respective variance inflation factors.

All analyses are conducted in the R programming language and environment [@rcoreteam2022]. For more details on these models and the cost-distance analysis, please see the supplementary materials.

{{< pagebreak >}}

# References

::: {#refs}
:::

### 2023-07-12 R/data_wrangling.R

@@ -1,0 +1,313 @@
+
+# Baja Ranching -- Data Wrangling
+
+##############################.
+#     CONTENTS
+# 01. Preamble
+# 02. Baja Califonia Sur
+# 03. Ranches
+# 04. Project Window
+# 05. Watersheds
+# 06. Springs
+# 07. Roads
+# 08. Digital Elevation Model (DEM)
+# 09. Cities
+##############################.
+
+# 01) PREAMBLE ------------------------------------------------------------
+
+# Load libraries
+library(dplyr)
+library(FedData)
+library(rgeoboundaries)
+library(here)
+library(lwgeom)
+library(osmdata)
+library(readr)
+library(sf)
+library(terra)
+
+gpkg <- here("data", "choyero.gpkg")
+
+# coordinate reference system
+# epsg: 4485 (Mexico ITRF92 / UTM zone 12N)
+# see https://epsg.io/4485
+
+# 02) BAJA CALIFORNIA SUR --------------------------------------------------
+
+mexico <- geoboundaries("mexico", "adm1") |> 
+  select(shapeName, shapeISO) |> 
+  rename_with(gsub, pattern = "shape", replacement = "") |> 
+  rename_with(tolower) |> 
+  st_transform(4485)
+
+bcs <- mexico |> filter(name == "Baja California Sur")
+
+write_sf(
+  mexico,
+  dsn = gpkg,
+  layer = "mexico"
+)
+
+write_sf(
+  bcs,
+  dsn = gpkg,
+  layer = "baja"
+)
+
+# 03) RANCHES -------------------------------------------------------------
+
+ranches <- here("data", "ranches.csv") |> 
+  read_csv() |>
+  # one ranch is missing coordinates
+  filter(!is.na(lat) | !is.na(long)) |> 
+  select(id, ranch, cluster, community, occupied, population, land, notes, lat, long) |> 
+  mutate(
+    community = case_when(
+      community == ''      ~ "Abandoned",
+      community == 'LH'    ~ "La Higuera",
+      community == 'LS'    ~ "La Soledad",
+      community == 'SMDT'  ~ "Santa Maria de Toris",
+      community == 'SPDLP' ~ "San Pedro de la Presa",
+      TRUE                 ~ community
+    )
+  ) |>
+  st_as_sf(coords = c("long", "lat"), crs = 4326) |>
+  st_transform(4485)
+
[…]

### 2023-07-12 R/overview-map.R

@@ -1,0 +1,229 @@
+
+# project area overview map
+
+# R preamble --------------------------------------------------------------
+
+library(here)
+library(ggrepel)
+library(ggspatial)
+library(patchwork)
+library(sf)
+library(tidyverse)
+library(viridis)
+
+# plot defaults -----------------------------------------------------------
+
+theme_set(theme_bw(12))
+
+theme_update(
+  axis.text = element_blank(),
+  axis.ticks = element_blank(),
+  axis.title = element_blank(),
+  legend.background = element_blank(),
+  legend.key.height = grid::unit(0.5, "cm"),
+  legend.spacing = unit(2, "pt"),
+  legend.text = element_text(margin = margin()),
+  legend.title = element_text(margin = margin()),
+  panel.background = element_rect(fill = "skyblue"),
+  panel.grid = element_blank(),
+  plot.margin = margin()
+)
+
+stroke <- 0.3
+
+# data --------------------------------------------------------------------
+
+gpkg <- here("data", "choyero.gpkg")
+
+mexico <- read_sf(gpkg, layer = "mexico")
+
+bcs <- read_sf(gpkg, layer = "baja")
+
+window <- read_sf(gpkg, layer = "window")
+
+cities <- read_sf(gpkg, layer = "cities")
+
+roads <- read_sf(gpkg, layer = "roads")
+
+highways <- read_sf(gpkg, layer = "highways")
+
+watersheds <- read_sf(gpkg, layer = "watersheds")
+
+ranches <- read_sf(gpkg, layer = "ranches")
+
+clusters <- read_sf(gpkg, layer = "clusters")
+
+# regional overview -------------------------------------------------------
+
+bb8 <- st_bbox(bcs) + c(-30000, -15000, 60000, 30000)
+
+region <- mexico |> 
+  filter(name != "Baja California Sur") |> 
+  st_transform(4485) |> 
+  st_intersection(st_as_sfc(bb8)) |> 
+  st_union()
+
+city_labels <- cities |> 
+  st_coordinates() |> 
+  as_tibble() |> 
+  bind_cols(st_drop_geometry(cities)) |> 
+  mutate(name = c("Ciudad Constitucion ", " La Paz"))
+
+overview <- ggplot() +
+  geom_sf(
+    data = region, 
+    fill = "gray95",
+    size = stroke
+  ) +
+  geom_sf(
[…]

### 2024-01-24 manuscript/reviews-response.md
reviews and response to reviews

We would like to thank the two reviewers and the editor for their time and
effort put into reviewing our manuscript. They offered a number of important
comments that we think make the paper better as a whole. We made substantial
revisions to the methods section as requested by reviewer 2. Reviewer 1 was more
concerned with conceptual and theoretical issues in the paper. Responding to
their concerns required some substantial changes to the structure of the paper,
including new sections in the intro, methods, and discussion.

## EDITOR

### MAJOR COMMENTS

1.  The abstract has a little shortfall compared with the richness of the body.
    Please especially enrich the methodology, result, and discussion parts.

    Response: The abstract was substantially revised to add requested detail and
    to reflect changes made in response to reviewer concerns.

2.  The current Introduction includes some component of methodology (GLM and
    model variables) as the reviewer #2 says. Please restructure the story line
    of Introduction by referring some comments of the reviewer #1. And in the
    last paragraph of Introduction please explicitly state the research
    questions, not methodological advantages in the current manuscript.

    Response: Discussion of methodology has been removed from the introduction
    and the primary research question restated at the end of the introduction
    for emphasis. Please see our comments below for how we address reviewer 1's
    concerns.

3.  The current result table1 is too small for readers. Please add the full
    statistical tables in the supplementary B 5.1 section. It is not so annoying
    but gives rich information to the readers. Additionally, It looks that the
    analysis of "land secure" factor disappeared somewhere in the manuscript.
    Please enhance the description of land fragmentation part.

    Response: Model results table was expanded to include all models.

4.  Both of the reviewers recommended to introduce additional references in
    Discussion part. Please recheck that Discussion contains the unconscious
    assumptions or conclusion without evidences.

    Response: Done. See below for more details about references we added here
    and throughout the paper.

### MINOR COMMENTS

1.  Can you show some photo of the landscape of the target regions to help the
    SUST readers understanding if possible?

    Response: Done!

2.  Please add some references of "Djikstra's algorithm" and "Tobler's Law" and
    other concepts without references if contained.

    Response: References added.

## REVIEWER 1

1.  Is the economy really subsistence based?

    Response: We were over-zealous in our description. They do produce a
    substantial portion of their own diet, but they are also not entirely
    dependent on what they produce, so it is probably more appropriate to refer
    to these communities as "smallholder" ranching economies. To address this,
    we have replaced "subsistence" with "smallholder" everywhere in the text. We
    also emphasize that they were never fully isolated from markets.

2.  Are all springs equally attractive to settlement?

    Response: Indeed, springs in this area are effectively the same. Lerback et
    al (2023) provides a fine-grained analysis of spring systems in the project
    area. They are primarily lentic pools that occur in the arroyos, so it's
    ideal to setup your ranch downslope of the spring on a flat piece of land
    slightly above the low point, so their house doesn't flood.

3.  Land fragmentation vs land insecurity.

    Response: The reviewer is right that these are not the same thing, so we
    have tried to make that more clear in the text. Basically, we see both land
    fragmentation and land insecurity as outcomes of the same process of
    privatization that was spawned by the Agrarian Land Reform of 1992 (in
    conjunction with NAFTA). There's a patchwork of property rights regimes out
    there that have consequences for land security/ownership. We define land
    (in)security in terms of whether individuals have explicit governmental
    documentation identifying rights to their land. Before the 1992 Land Reforms
    and NAFTA, land title to a specific piece of land wasn't really an issue
    because the land was treated as one contiguous commons co-managed by all the
    individual families (who were by and large related to one another through
    blood and marriage). As a result, very few individuals obtained
    documentation to land or became ejido members. After the 1992 Land Reforms,
    however, there were monetary incentives for ejido members to split up and
    sell land. This then raised issues for families that either never had clear
    title to land or never became members of the ejido. Once the land became
    parcelized and placed on land markets, individuals and organizations from
    outside the community began purchasing land (and sometimes fencing it off),
    creating an existential threat for land dis-appropriation. As a result, land
    fragmentation and land security are heavily related to one another today.

4.  Move "historical antecedents" to introduction.

    Response: As requested, additional historical context was added to the
    introduction. We also added a site content section to the methods to
    emphasize aspects of the region most relevant to the analysis.

5.  What about road condition and transport efficiency?

    Response: Road condition is incorporated into the cost-distance analysis
    under the assumption that travel is faster over maintained roads. That said,
    there's only really one way in or out of each watershed (north and south),
    so it's really how close they are to the maintained road within the study
    area that makes the biggest difference.

6.  Is the ecological discussion relevant?

    Response: Our goal is to show that these ranchers must navigate a trade-off
    between distant market services and local ecosystem services, and even as
    they integrate more with markets, they are still constrained by local
    ecology, which puts them in a precarious position. So, understanding the
    ecology, how it shapes their decisions and is shaped by their decisions is
    critically important. We have to tried to make the emphasis on this
    trade-off clearer by changing the title of the paper. It is also emphasized
    more in the abstract and in the introduction.

7.  What about intermediate markets?

    Response: As far as we know, there are no intermediate markets between these
    communities and the major population centers of Ciudad Constitucion and La
    Paz.

## REVIEWER 2

1.  Why binomial rather than logit GLM?

    Response: In R, the default link for the binomial GLM is the logit link. We
    should have been more clear about this, so we added detail to the methods to
    clarify.

2.  Model specification and functional form should be provided in the text.

    Response: The reviewer is absolutely right to request this, so we have added
    detail to the methods section to elaborate on this.

3.  Provide a table that defines the variables used along with some of their
    descriptive statistics.

    Response: This is a solid suggestion, so the table has been added.

4.  Land fragmentation vs land insecurity.

    Response: see response to reviewer 1.

5.  Assumptions of travel time analysis need citations.

    Response: The reviewer requested examples of the use of travel time in
    ethnographic and related studies involving market integration. We actually
    provided some of these, but they were not in the methods section on travel
    time estimation, which was probably a mistake on our part, so we have moved
    them there. We also added some additional context, emphasizing that our use
    of travel time estimates to various locations is a standard procedure in
    land use and transport interaction modeling in urban and economic geography,
    where studies commonly model residential choice behavior in a more urban
    context.

6.  Not clear how results of statistical analysis justify conclusion.

    Response: In the results section, we note the following. The binomial GLM is
    a model of ranch locations. The Poisson GLM is a model of ranch size. The
    binomial GLM suggests that ranches are more likely to be occupied the closer
    they are to springs. The exact opposite is true of distance to La Paz. They
    are more likely to be occupied farther from the city. This suggests that
    access to water rather than markets is a bigger constraint on settlement
    locations. But when you look at the size of ranches, the Poisson GLM says
    they are bigger the closer you get to cities and the farther you get from
    springs. So, in the discussion section, we summarize these ideas by noting
    that access to water explains where occupied ranches are located, and access
    to markets explains how big they are.

7.  Increasing wages in the cities may lead rural ranchers to shift time and
    energy away from ranching, decreasing their impacts on local ecology.

    Response: While we don't have data on how much of their income comes from
    cities, we do emphasize the role that demography plays in economics:
    population increases, this leads to increased demand for goods and services,
    and that in turn leads to increased demand for workers to supply those goods
    and services (demand for labor is synonymous with wages). This should
    incentivize movement toward the city, so we should get further levels of
    population decline in these ranching communities. We have added some text to
    emphasize that the consequence of this will be complex and multifaceted.

### 2024-01-24 R/least-cost-paths.qmd
re-run all code for analysis

@@
 library(here)
-library(hiker)
+library(igraph)
 library(sf)
@@
 # ORIGIN ---
-clusters <- 
-  read_sf(gpkg, layer = "clusters") |>
+clusters <- gpkg |> 
+  read_sf(layer = "clusters") |>
   mutate(variable = "cluster")
@@
 # DESTINATION --- to city, terminus at edge of window
-terminus <- 
-  read_sf(gpkg, layer = "terminus") |>
+terminus <- gpkg |> 
+  read_sf(layer = "terminus") |>
   mutate(variable = "terminus")
 
-springs <- 
-  read_sf(gpkg, layer = "springs") |>
+springs <- gpkg |> 
+  read_sf(layer = "springs") |>
   select(id, name) |>
@@
   filename = here("figures", "od-points.png"),
-  dpi = 600,
+  dpi = 800,
   width = 3.5,
@@
 
+The necessary functions are defined in the folded code chunk below.
+
+```{r}
+#| code-fold: true
+
+#' Cost surface
+#'
+#' Create a `terrain` representing the cost of travel. 
+#'
+#' The `SpatRaster` must have a projected CRS, with both distance units and elevation
+#' values in meters.
+#'
+#' @param x a `SpatRaster` with elevation values.
+#' @param hf a character string specifying the preferred hiking function. Options
+#'   include `tobler` and `campbell` (default).
+#' @param max_slope a numeric value specifying the maximum allowed hiking slope
+#'   _in degrees_ (default is 45). Suppose you ask: Would a person really hike
+#'   _that_? This is the shallowest slope where the answer to that question is No.
+#' @param neighbors numeric, one of 4, 8, or 16 (default) specifying the number of
+#'   adjacent cells between which to calculate movement cost. These are passed on
+#'   to `terra::adjacent()` and represent standard neighborhoods (rook, queen, and
+#'   knight + queen).
+#' @param ... Arguments passed on to the hiking function specified by `hf`.
+#'
+#' @return a `terrain` representing cost of travel.
+hf_terrain <- function(x,
+                       hf = "campbell",
+                       max_slope = 45,
+                       neighbors = 8,
+                       ...) {
+
+  # find adjacent cells in the raster
+  # make sure to ignore all na values
+  n_cells <- terra::ncell(x)
+
+  na_cells <- which(is.na(terra::values(x)))
+
+  cells <- setdiff(1:n_cells, na_cells)
+
+  adj <- terra::adjacent(
+    x,
+    cells = cells,
+    directions = neighbors,
+    pairs = TRUE
+  )
+
+  adj <- adj[!adj[, 2] %in% na_cells, ]
+
+  ### slope ###
+  # change in y between adjacent pixels
+  heights <- terra::values(x)[, 1]
+
+  rise <- (heights[adj[, 2]] - heights[adj[, 1]])
+
+  # change in x between adjacent pixels
+  run <- terra::distance(
+    terra::xyFromCell(x, adj[, 1]),
+    terra::xyFromCell(x, adj[, 2]),
+    lonlat = FALSE,
+    pairwise = TRUE
+  )
+
+  slope <- rise/run
+
+  ### speed ###
+  speed_fun <- switch(
+    hf,
+    "campbell" = campbell,
+    "tobler"   = tobler,
+    stop(paste0("hiker does not implement the ",
+                hf,
+                " hiking function."))
+  )
+
+  speed <- speed_fun(slope, ...)
+
+  ### conductance ###
+  conductance <- speed/run
+
+  # what is a realistic slope people would try to hike?
+  max_slope <- tan(max_slope * pi / 180) # degrees -> radians -> rise-over-run
+
+  i <- which(abs(slope) >= max_slope)
+
+  conductance[i] <- 0
+
+  # build matrix
+  cm <- Matrix::spMatrix(nrow = n_cells, ncol = n_cells)
+
+  cm[adj] <- conductance
+
+  # include some spatial and other information to make
+  # manipulating the network faster and easier
+  # when passed to other functions
+  # storing bounding box as vector to keep the size down
+  bb8 <- terra::ext(x)
+  bb8 <- c(bb8$xmin, bb8$xmax, bb8$ymin, bb8$ymax)
+
+  xcrs <- terra::crs(x, describe = TRUE)
+  xcrs <- paste0(xcrs$authority, ":", xcrs$code)
+
+  tmin <- 1/max(conductance)
+  tmax <- 1/min(conductance[conductance > 0])
+
+  terrain <-
+    list(
+      "conductance" = cm,
+      "neighbors"   = neighbors,
+      "nrow"        = terra::nrow(x),
+      "ncol"        = terra::ncol(x),
+      "bb8"         = bb8,
+      "crs"         = xcrs,
+      "range"       = c("min" = tmin, "max" = tmax)
+    )
+
+  class(terrain) <- "terrain"
+
+  return(terrain)
+
+}
+
+campbell <- function(x, decile = 50L) {
+
+  if (length(decile) != 1) {
+
+    stop("campbell accepts only one decile at a time.")
+
+  }
+
+  # campbell function wants degrees, not rise-over-run as in tobler
+  # but, we're working with rr, so remember to convert radians to degrees with 180/pi!
+  slope <- atan(x) * 180 / pi
+
+  # values provided in Campbell 2019, Supplement, Table S1
+  # for simplicity, I excluded all but the deciles
+  deciles <-
+    data.frame(
+      decile = c(10L, 20L, 30L, 40L, 50L, 60L, 70L, 80L, 90L),
+      a = c(-1.568, -1.71,  -1.858, -1.958, -2.171, -2.459, -2.823,  -3.371,  -3.06),
+      b = c(13.328, 10.154,  8.412,  8.96,  10.064, 11.311, 12.784,  15.395,  16.653),
+      c = c(38.892, 36.905, 39.994, 50.34,  63.66,  79.287, 98.697, 134.409, 138.875),
+      d = c( 0.404,  0.557,  0.645,  0.649,  0.628,  0.599,  0.566,   0.443,   0.823),
+      e = c(-0.003, -0.004, -0.004, -0.005, -0.005, -0.005, -0.005,  -0.005,  -0.0139)
+    )
+
+  ind <- which(deciles$decile == as.integer(decile))
+
+  deciles <- deciles[ind, ]
+
+  a <- deciles$a
+  b <- deciles$b
+  c <- deciles$c
+  d <- deciles$d
+  e <- deciles$e
+
+  # lorentz distribution
+  lorentz <- (1 / ((pi * b) * (1 + ((slope - a) / b)^2)))
+
+  # modified lorentz
+  (c * lorentz) + d + (e * slope)
+
+}
+
+#' Channel of reduced cost
+#'
+#' Reduces cost of travel along paths that cross through channels or
+#' other aspects of a terrain that expedite travel. 
+#'
+#' @param x a cost `terrain` generated by `hf_terrain()`.
+#' @param channel an `sf` object.
+#' @param .m a numeric between 0 and 1 applied as a weight to the cost of travel.
+#'
+#' @return a `terrain` representing cost of travel.
+hf_channel <- function(x, channel, .m = 1) {
+
+  rr <- terra::rast(
+    nrow   = x$nrow,
+    ncol   = x$ncol,
+    extent = terra::ext(x$bb8),
+    crs    = x$crs
+  )
+
+  channel <- terra::vect(channel)
+
+  cells <- terra::cells(rr, channel)[, 2]
+
+  adj <- cbind(
+    "from" = as.integer(x$conductance@i + 1),
+    "to"   = as.integer(x$conductance@j + 1)
+  )
+
+  # find incident edges, from or to cells intersecting barrier
+  i <- which(adj[, 1] %in% cells | adj[, 2] %in% cells)
+  adj <- adj[i, ]
+
+  # want to reduce time, which means dividing conductance by .m
+  x$conductance[adj] <- (x$conductance[adj] / .m)
+
+  return(x)
+
+}
+
+```
+
 ```{r}
@@
 #| code-fold: true
+
+#' Terrain to raster
+#'
+#' Converts a cost `terrain` into a raster representing
+#' travel cost, specifically the average cost of moving from/to each
+#' cell. Mostly useful for visualizing with `terra::plot()`
+#' as a sanity check.
+#'
+#' @param x a cost `terrain` generated by `hf_terrain()`.
+#'
+#' @return a `SpatRaster` with travel cost values.
+hf_rasterize <- function(x) {
+
+  rr <- terra::rast(
+    nrow   = x$nrow,
+    ncol   = x$ncol,
+    extent = terra::ext(x$bb8),
+    crs    = x$crs
+  )
+
+  # this part is basically what Jacob van Etten does with
+  # gdistance::raster(TransitionLayer), though that offers more options
+  # here I am just averaging the vertical and horizontal sums
+  total_row <- Matrix::rowSums(x$conductance)
+  total_col <- Matrix::colSums(x$conductance)
+
+  #logical sparse matrix, identifies number of adjacent cells
+  lm <- methods::as(x$conductance, "lMatrix")
+
+  n_row <- Matrix::rowSums(lm)
+  n_col <- Matrix::colSums(lm)
+
+  # average cost of moving from/to each cell
+  # values <- total_cost / n_adjacent
+  v_row <- total_row / n_row
+  v_col <- total_col / n_col
+
+  values <- (v_row + v_col)/2
+
+  terra::setValues(rr, (1/values))
+
+}
 
@@
 ### To springs
+
+Function to calculate paths and travel times is in the folded code chunk below.
+
+```{r}
+#| code-fold: true
+
+#' Shortest path between points
+#'
+#' Calculate the "shortest" or least-cost paths between multiple start and end points.
+#' For each from-point, the shortest path to every to-point is calculated.
+#'
+#' @param x a cost `terrain` generated by `hf_terrain()`.
+#' @param from an `sf` specifying start locations, must be POINT or MULTIPOINT.
+#' @param to an `sf` specifying end locations, must be POINT or MULTIPOINT.
+#' @param add_cost logical, whether to include a column of travel costs.
+#'
+#' @return An `sf` object with least-cost `LINE` features and two columns: from and to.
+#'   Values in the from- and to-columns are `1:nrow(from)` and `1:nrow(to)`, respectively.
+#'   If `add_cost = TRUE`, the `sf` object also includes a travel cost column.
+hf_hike <- function(x, from, to, add_cost = FALSE) {
+
+  from_xy <- sf::st_coordinates(from)[, 1:2, drop = FALSE]
+  to_xy   <- sf::st_coordinates(to)[, 1:2, drop = FALSE]
+
+  rr <- terra::rast(
+    nrow   = x$nrow,
+    ncol   = x$ncol,
+    extent = terra::ext(x$bb8),
+    crs    = x$crs
+  )
+
+  from_cells <- terra::cellFromXY(rr, from_xy)
+  to_cells   <- terra::cellFromXY(rr, to_xy)
+
+  graph <- igraph::graph_from_adjacency_matrix(
+    x$conductance,
+    mode = "directed",
+    weighted = TRUE
+  )
+
+  # invert conductance to get travel cost
+  igraph::E(graph)$weight <- (1 / igraph::E(graph)$weight)
+
+  # return a list of lists of vectors of vertices on paths
+  shorties <-
+    lapply(
+      from_cells,
+      function(z) {
+
+        res <-
+          igraph::shortest_paths(
+            graph,
+            from = z,
+            to = to_cells,
+            mode = "out"
+          )
+
+        res$vpath
+
+      })
+
+  # collapse (list of lists of vectors) to (list of vectors)
+  shorties <- unlist(shorties, recursive = FALSE)
+
+  # now we can just make a bunch of linestrings out of them
+  line_list <-
+    lapply(
+      shorties,
+      function(z) {
+
+        cells <- as.vector(z)
+
+        # to handle case where from-vertex == to-vertex
+        if (length(cells) == 1) cells <- rep(cells, 2)
+
+        xy <- terra::xyFromCell(rr, cells)
+
+        sf::st_linestring(xy)
+
+      }
+    )
+
+  # and then an sf
+  sf_col <- sf::st_sfc(line_list, crs = x$crs)
+
+  short_paths <- sf::st_sf(geometry = sf_col)
+
+  # add indices
+  short_paths <- transform(
+    short_paths,
+    from = rep(1:nrow(from_xy), each = nrow(to_xy)),
+    to   = rep(1:nrow(to_xy), times = nrow(from_xy))
+  )
+
+  if (add_cost) {
+
+    cost <- igraph::distances(graph, from_cells, to_cells, mode = "out")
+
+    short_paths <- transform(short_paths, cost = c(t(cost)))
+
+  }
+
+  return(short_paths)
+
+}
+
+```
 
@@
 
+od_points <- bind_rows(
+  od_points |> mutate(destination = ifelse(variable == "spring", "To springs", "To cities")),
+  od_points |> filter(variable == "cluster") |> mutate(destination = "To springs")
+)
+
 paths <- bind_rows(
@@
   ) +
-  facet_wrap(vars(destination)) +
   geom_sf(
@@
     stroke = 0.3,
-    size = ifelse(od_points$variable == "city", 2.2, 1.5) |> rep(2)
+    size = ifelse(od_points$variable == "city", 2.2, 1.5)
   ) +
+  facet_wrap(vars(destination)) +
   scale_color_manual(values = c("grey15", "white", "white")) +

### 2024-01-24 R/models.qmd
re-run all code for analysis

@@
 library(car)
-library(dplyr)
-library(ggplot2)
 library(here)
@@
 library(patchwork)
-library(purrr)
 library(sf)
 library(sfdep)
-library(tidyr)
+library(terra)
+library(tidyterra)
+library(tidyverse)
 library(viridis)
@@
       row.striping.include_stub = FALSE,
-      row.striping.include_table_body = TRUE,
+      # row.striping.include_table_body = TRUE,
       table.align = "left",
@@
 remove(path_to_cities, path_to_springs, security, ranches)
+
+```
+
+### Topography
+
+For additional context, here's what the terrain looks like.
+
+```{r}
+
+elevation <- here("data", "dem_30m.tif") |> rast()
+
+elevation_stats <- global(
+  elevation, 
+  fun = c("min", "max", "mean", "sd"), 
+  na.rm = TRUE
+)
+
+```
+
+```{r}
+#| code-fold: true
+
+elevation_stats |> 
+  pivot_longer(
+    everything(), 
+    names_to = "statistic",
+    values_to = "value"
+  ) |> 
+  gt() |> 
+  tab_header("Elevation") |> 
+  fmt_number(-statistic, decimals = 1) |> 
+  cols_width(statistic ~ px(125)) |> 
+  tab_options(
+    column_labels.hidden = TRUE,
+    data_row.padding = px(6),
+    heading.align = "left",
+    heading.border.bottom.style = "hidden",
+    row.striping.background_color = "gray96",
+    row.striping.include_stub = FALSE,
+    row.striping.include_table_body = TRUE,
+    table.align = "left",
+    table.border.top.style = "hidden",
+    table.border.bottom.style = "hidden"
+  )
+
+```
+
+```{r}
+#| code-fold: true 
+#| fig-width: 6.5
+#| out-width: "50%"
+
+hill <- shade(
+  terrain(elevation, "slope", unit = "radians"), 
+  terrain(elevation, "aspect", unit = "radians"), 
+  angle = 30, 
+  direction = seq(45, 360, by = 45)
+)
+
+hill <- setValues(hill, scales::rescale(values(hill), to = c(1, 1000)))
+hill <- round(hill)
+
+ggplot() +
+  geom_spatraster(
+    data = hill, 
+    fill = hcl.colors(1000, "Grays")[values(hill)],
+    alpha = 0.75,
+    maxcell = Inf
+  ) +
+  geom_spatraster(data = elevation, alpha = 0.5) +
+  scale_fill_hypso_c(
+    name = "Elevation (m)",
+    palette = "utah_1",
+    limits = elevation_stats[, c("min", "max")] |> t() |> c(),
+    alpha = 0.5,
+    na.value = "transparent"
+  ) +
+  theme(
+    axis.text = element_blank(),
+    axis.ticks = element_blank(),
+    axis.title = element_blank(),
+    legend.position = "none"
+  )
+
+```
+
+### Data Summary
+
+```{r}
+
+data_summary <- function(x, .m, .v){
+
+  .f <- list(min = min, max = max, mean = mean, sd = sd)
+
+  x |> 
+    st_drop_geometry() |> 
+    mutate(model = .m) |> 
+    group_by(model) |> 
+    summarize(across({{ .v }}, .fns = .f)) |> 
+    pivot_longer(
+      -model, 
+      names_sep = "_", 
+      names_to = c("variable", "statistic")
+    ) |> 
+    pivot_wider(
+      names_from = statistic,
+      values_from = value
+    )
+
+}
+
+variable_summary <- bind_rows(
+  clusters |> data_summary("Binomial", c(occupied, springs, paz)),
+  clusters |> filter(population > 0) |> data_summary("Poisson", c(population, springs, cities, secure))
+)
+
+remove(data_summary)
+
+write_csv(variable_summary, here("data", "variable-summary.csv"))
+
+variable_summary |> 
+  group_by(model) |> 
+  gt() |> 
+  fmt_number(decimals = 2) |> 
+  tab_options(
+      column_labels.font.weight = "bold",
+      data_row.padding = px(6),
+      heading.align = "left",
+      heading.border.bottom.style = "hidden",
+      row_group.as_column = TRUE,
+      table.border.top.style = "hidden",
+      table.border.bottom.style = "hidden"
+    ) |> 
+    sub_missing(missing_text = "")
 
@@
 vif <- models |> 
-  map(vif) |> 
+  map(car::vif) |> 
   bind_rows(.id = "model")
@@
     subtitle = "Alternative: difference in log-likelihood greater than zero"
-  )
+  ) |> 
+  cols_width(model ~ pct(21.5))
 
@@
   tab_header(
-    title = "Variance Inflation",
-    subtitle = "Alternative: "
-  )
+    title = "Variance Inflation"
+  ) |> 
+  cols_width(model ~ pct(21.5))
 

### 2024-01-24 R/overview-map.R

@@ -29,8 +29,6 @@ theme_update(
   plot.margin = margin()
 )
 
-stroke <- 0.3
-
 # data --------------------------------------------------------------------
 
 gpkg <- here("data", "choyero.gpkg")
@@ -73,12 +71,12 @@ overview <- ggplot() +
   geom_sf(
     data = region, 
     fill = "gray95",
-    size = stroke
+    size = 0.3
   ) +
   geom_sf(
     data = bcs, 
     fill = "#A8A076",
-    size = stroke
+    size = 0.3
   ) +
   geom_sf(
     data = highways |> st_buffer(1750) |> st_union(), 
@@ -97,8 +95,7 @@ overview <- ggplot() +
     aes(X, Y, label = name),
     size = 10/.pt,
     nudge_x = c(-100000, 70000),
-    nudge_y = 80000,
-    hjust = c(1,0)
+    nudge_y = 80000
   ) +
   geom_sf(
     data = cities, 
@@ -106,7 +103,7 @@ overview <- ggplot() +
     fill = "#fffbc9", 
     color = "black",
     size = 3,
-    stroke = stroke
+    stroke = 0.3
   ) +
   annotation_scale(
     aes(location = "bl"),
@@ -131,17 +128,18 @@ base <- ggplot() +
   geom_sf(
     data = region, 
     fill = "#A8A076",
-    size = stroke
+    size = 0.3
   ) +
   geom_sf(
     data = watersheds,
     fill = alpha("gray95", alpha = 0.5), 
     color = "gray50",
-    size = stroke
+    size = 0.3
   ) +
   geom_sf(
     data = roads |> filter(level != "T"),
-    color = "white"
+    color = "white",
+    linewidth = 0.1
   )
 
 # population map ----------------------------------------------------------
@@ -223,7 +221,7 @@ population + overview + land + guide_area() + plot_layout(guides = 'collect')
 
 ggsave(
   here("figures", "overview-map.png"),
-  dpi = 600,
+  dpi = 800,
   width = 5.75,
   height = 5.75
 )

### 2024-01-24 manuscript/manuscript.qmd
changes to manuscript requested by reviewer

@@
 ---
-title: Market integration mediates human-environment interactions among traditional ranchers in Baja California Sur, Mexico
+title: Trade-off between market and ecosystem services drives settlement decisions among smallholder ranchers in Baja California Sur, Mexico
 date: today
@@
 target-journal:
-  - "Nature Sustainability"
-  - "Human Ecology"
-  - "PNAS"
+  - "Sustainability Science"
 abstract: |
-  Market integration presents new opportunities and challenges for traditional communities still engaged in a largely subsistence-based economy.
-In this paper, we explore one particular aspect of that dynamic: how market integration affects the way that populations of historically marginalized ranchers distribute themselves across a landscape, where they feel the competing pull of urban amenities and local ecosystem services.
-Our study focuses on a small community of ranchers living in kin-based ranch clusters in the rugged and arid landscapes of the Sierra de la Giganta mountains along the eastern spine of the Baja California peninsula in Mexico.
+  While smallholder food producers increasingly depend on goods and services provided by distant markets, they are still constrained by their local ecology, as their rural communities often lack many of the infrastructural advantages of urban centers.
+Navigating the costs and benefits of market and ecosystem services is, as a consequence, crucial to their livelihoods and overall well-being.
+Here, we explore one aspect of that market-ecology trade-off: its effect on the residential choice behavior of ranchers living in the arid Sierra de la Giganta mountains along the eastern spine of the Baja California peninsula in Mexico.
+We rely on two Generalized Linear Models (GLMs), a binomial GLM of the locations of occupied and abandoned ranches and a Poisson GLM of population size at occupied ranches.
+Our proxies for market integration and ecosystem services are travel-time estimates to surficial springs and distant cities.
+By using these travel time estimates, we enhance the ethnographic study of residential choice behavior in smallholder systems by drawing on key innovations from urban and economic geography.
 Our models show that ranch clusters are more likely to be abandoned the farther they are from surficial springs, and occupied ranch clusters have larger populations the closer they are to markets.
-This has important consequences for the ecological resilience of this ranching community, particularly in the context of climate change.
+This has important consequences for the ecological resilience of this ranching community, particularly in the context of climate-induced drought, which may lead to ranching strategies that are more locally resilient but less profitable on the market.
+At the same time, market integration may reshape the social structure of these communities, introducing new incentives that lead them to further integrate with and increasingly depend on those markets.
+All of these changes are exacerbated by demographic processes, in particular the explosive growth of the region's urban centers.
 keywords:
-  - Market Integration
-  - Ideal Free Distribution
-  - Behavioral Ecology
+  - Habitat Suitability
+  - Land Use 
+  - Transportation
   - Sustainability
@@
 linenumbers: false
-watermark: false
 bibliography: bibliography.bib
@@
   simple-pdf: default
+  docx: default
 execute: 
@@
 library(knitr)
+library(readr)
 
@@
 
-In traditional, small-scale subsistence economies, the value of a habitat is typically a function of its local resources, particularly their distribution and abundance.
-As individuals integrate with modern markets, however, that connection becomes more tenuous.
-This can be a good thing, of course, allowing individuals to diversify their livelihood, making them more flexible and capable of absorbing potential risks and uncertainty [@barrett2001; @little2001].
-But, it can have other, unforeseen consequences, too, especially as individuals come to depend on markets for their livelihood.
-For those living in predominantly marginal habitats, it may even have the opposite effect, making them more sensitive to the effects of climate change and other exogenous forces like infrastructure development and range fragmentation [@fratkin2001; @hobbs2008; @thornton2009].
-Together, these conflicting outcomes suggest important trade-offs for individuals presented with the opportunity to integrate more closely with modern markets.
-Time and energy they put into accessing regional market services is often time and energy they cannot spend accessing local ecosystem services.
-How they navigate that trade-off can have important consequences for their health and well-being [see, for example, @liebert2013; @urlacher2016; @stagaman2018], including dramatic changes to their life history and demography [@shenk2016; @colleran2015].
-It is, thus, critically important to understand how and to what degree individuals choose to integrate with modern markets.
+In traditional, small-scale subsistence economies, the value of a habitat is
+typically a function of its local resources, particularly their distribution and
+abundance [@moran2022]. As individuals integrate with modern markets, however,
+that connection becomes more tenuous. This can be a good thing, of course,
+allowing individuals to diversify their livelihood, making them more flexible
+and capable of absorbing potential risks and uncertainty [@barrett2001;
+@little2001]. But, it can have other, unforeseen consequences, too, especially
+as individuals come to depend on markets for their livelihood. For those living
+in predominantly marginal habitats, it may even have the opposite effect, making
+them more sensitive to the effects of climate change and other exogenous forces
+like infrastructure development, range fragmentation, and land insecurity
+[@fratkin2001; @hobbs2008; @thornton2009]. Together, these conflicting outcomes
+suggest important trade-offs for individuals presented with the opportunity to
+integrate more closely with modern markets. Time and energy they put into
+accessing regional market services is often time and energy they cannot spend
+accessing local ecosystem services. How they navigate that trade-off can have
+important consequences for their health and well-being [see, for example,
+@liebert2013; @urlacher2016; @stagaman2018], including dramatic changes to their
+life history and demography [@shenk2016; @colleran2015]. It is, thus, critically
+important to understand how and to what degree individuals choose to integrate
+with modern markets.
 
-Here we explore that question, focusing on a small group of historically marginalized rural ranchers living in the arid Sierra de la Giganta region of Baja California Sur, Mexico [@macfarlan2020a, see @fig-overview-map].
-While these people are sometimes referred to as "choyeros", a local nickname for anyone born in the rural, mountainous ranching communities of Baja California Sur (also a reference to the local cholla cactus, *Cylindropuntia cholla*), in this paper, we refer to them simply as "ranchers" or "rancheros", which are the expressions they typically use to describe themselves.
-Descended from Euro-American colonists who began settling the region in the late-17th century [@crosby1994], they number approximately 4,000 today, living in communities largely organized around clusters of one or more ranches housing inter-generational kin.
-These communities are some of the most severely under-served in the region, lacking access to critical services and infrastructure, including piped water, sanitation, and energy [@secretaríadedesarrollosocial2021].
+To explore this issue, we focus on a small group of historically marginalized
+rural ranchers living in the arid Sierra de la Giganta region of Baja California
+Sur (BCS), Mexico [@macfarlan2020a, see @fig-overview-map]. We first explore how
+market integration and local ecology structure the distribution of occupied and
+abandoned ranch clusters. This is a process that took place over several hundred
+years, with the shift to its current distribution beginning in the late 1970s,
+following construction of the transpeninsular Federal Highway 1 [@barkenbus1974;
+@mathes2015]. We then ask how market integration and local ecology have shaped
+the distribution of the population across occupied clusters today.
 
-Given the aridity of the area and lack of regionally connective water infrastructure, rancheros depend almost entirely on surficial springs to satisfy their water needs, including drinking water for themselves and their livestock, as well as irrigation for garden plots (known as huertas and jardines), which provide a small supplement to human and livestock diets during dry periods [@macfarlan2020a; @macfarlan2020b; @schniter2021; @lerback2022].
-To make use of springs, ranchers either have to pump water themselves, using gas or solar powered equipment, or transport it by gravity irrigation [@lerback2022].
-A strong incentive thus exists to locate in closer proximity to these water resources.
+## Historical Context
 
-A similar, though competing, incentive exists for accessing markets in urban areas like Loreto, Ciudad Constitución, and La Paz, the capital of Baja California Sur.
-These cities furnish rural ranchers with many of the necessities they cannot acquire locally, including the ability to trade livestock, register animals with the state, receive medical attention, renew water permits, purchase groceries, and socialize with people outside their local community.
-The chance for supplemental income will also on occasion draw rancheros to the cities, as it has for pastoralists elsewhere [@barrett2001; @little2001].
-This includes working in construction, seasonal employment in coastal fisheries, or the tourist sector, specifically within the resort towns of Cabo San Lucas and San Jose del Cabo.
+While these people are sometimes referred to as "choyeros", a local nickname for
+anyone born in the rural, mountainous ranching communities of BCS (also a
+reference to the local cholla cactus, *Cylindropuntia cholla*), in this paper,
+we refer to them simply as "ranchers" or "rancheros", which are the expressions
+they typically use to describe themselves. Descended from Euro-American
+colonists who began settling the region in the late-17th century [@crosby1994],
+they number approximately 4,000 today, living in communities largely organized
+around clusters of one or more ranches housing inter-generational kin
+[@macfarlan_etal2020a].
 
-As a baseline for investigating the demographic consequences of these competing necessities, we first explore how market integration and local ecology structure the distribution of occupied and abandoned ranch clusters.
-This is a process that took place over several hundred years, with the shift to its current distribution beginning in the late 1970s, following construction of the transpeninsular Federal Highway 1 [@barkenbus1974].
-We then ask how market integration and local ecology have shaped the distribution of the population across occupied clusters today.
+Historically, the isolation of these rural ranching communities in the
+mountainous hinterlands of the Sierra de la Giganta meant that getting to the
+markets found in nearby urban centers (mostly in the capital of La Paz) was an
+extremely costly endeavor, even more so than it is today. This had two important
+consequences for these rural ranchers. First, while they do rely on markets for
+many of their subsistence needs, local food production still contributes to a
+substantial portion of their diet [@macfarlan_etal2023a], meaning they are
+neither fully subsistence-based nor fully market integrated. For this reason,
+they are more accurately characterized as smallholder ranchers rather than pure
+subsistence ranchers or pastoralists [@ifad2013]. Second, their communities
+remain some of the most severely under-served in the region, lacking access to
+critical services and infrastructure, including piped water, sanitation, and
+energy [@secretaríadedesarrollosocial2021].
+
+Given the aridity of the area and lack of regionally connective water
+infrastructure, rancheros depend almost entirely on surficial springs to satisfy
+their water needs, including drinking water for themselves and their livestock,
+as well as irrigation for garden plots (known as huertas and jardines), which
+provide a small supplement to human and livestock diets during dry periods
+[@macfarlan2020a; @macfarlan_etal2020a; @schniter2021; @lerback2022]. To make
+use of springs, ranchers either have to pump water themselves, using gas or
+solar powered equipment, or transport it by gravity irrigation [@lerback2022]. A
+strong incentive thus exists to locate in closer proximity to these water
+resources.
+
+A similar, though competing, incentive exists for accessing markets in urban
+areas like Loreto, Ciudad Constitución, and La Paz, the capital of BCS. These
+cities furnish rural ranchers with many of the necessities they cannot acquire
+locally, including the ability to trade livestock, register animals with the
+state, receive medical attention, renew water permits, purchase groceries, and
+socialize with people outside their local community. The chance for supplemental
+income will also on occasion draw rancheros to the cities, as it has for
+pastoralists elsewhere [@barrett2001; @little2001]. This includes working in
+construction, seasonal employment in coastal fisheries, or the tourist sector,
+specifically within the resort towns of Cabo San Lucas and San Jose del Cabo.
+
+Complicating matters is the fact that land privatization serves as an effective
+constraint on these dynamics. This is important because pastoralists have
+historically lacked clear title to land, instead relying on open rangeland
+systems [@behnke1994; @turner1999]. In Mexico, these systems take the form of
+*ejidos*, or areas of communal agricultural land. The first *ejidos* were
+established by the Mexican state in the early part of the twentieth century
+[@perramond2008], though in our project area they did not arrive until 1975
+[@dof1975] and 1976 [@dof1976], respectively. While these *ejidos* have
+historically served as a buffer against land fragmentation, allowing rancheros
+to maintain their way of life, their ability to play that role continues to
+change as the pace of privatization and shifting land tenure accelerates
+[@lesorogol2003; @reid_etal2008].
+
+In BCS (along with the rest of Mexico), the Agrarian Land Reforms of 1992 played
+an outsized role in driving these changes, as they introduced powerful new
+incentives for *ejidos* to parcelize collective land holds and allocate private
+rights to individuals [@barnes2009]. While note the sole cause, this new
+property regime has increased uncertainty and insecurity about land title and
+exposed the region to speculation from real estate markets. As has happened
+elsewhere, these privatization efforts appear to be led by NGOs, commercial
+interests, and other external forces, with the result being greater
+fragmentation of pastoral rangeland systems and increased uncertainty about land
+use [@burnsilver_etal2008; @galvin2009]. For the rural ranchers of the Sierra de
+la Giganta, this process of fragmentation and increasing insecurity at least
+partially explains why ranches of related individuals tend to cluster.
+Rancheros, unable to find new land for themselves, simply build ranches next to
+their kin on land that is at least historically associated with their family.
+
+To summarize, we want to know how traditional rural ranchers in BCS navigate the
+competing costs and benefits of market integration and local ecosystem services.
+This is done by estimating their contribution within a regression framework to
+the location and size of ranch clusters in two east-west running watersheds
+along the southern extent of the Sierra de la Giganta range. Given the extreme
+ecological constraints that characterize this region, especially the aridity of
+the region, we expect water management to play a substantial role in individual
+settlement decisions. Privatization may impose an additional constraint on this
+system, especially as it has served to increase land insecurity.
+
+# Methods
+
+## Study Area
+
+As shown in @fig-overview-map, our study area encompasses two east-west running
+watersheds with a combined area of roughly 1,042 km$^2$. The area is
+approximately equidistant in Cartesian terms between the two nearest urban
+centers, being approximately 85 km from Ciudad Constitucion and 95 km from La
+Paz. This is an extremely rugged and mountainous terrain (see
+@fig-context-photos C), as the elevation of this area ranges from 0 to 980 m,
+with a mean of 427 m and a variance of 142 m. The average slope is around 15°.
+For reference, just 35 km southwest of the study region, between it and Federal
+Highway 1, the average slope over a comparable area is just 0.8°.
+
+As mentioned above, this is also an extremely arid region. It is a southern
+extension of the Sonoran Desert Biogeographic Region, so the area is dominated
+by columnar cacti and xeric scrub [@shreve_wiggins1964;
+@leondelaluz_dominguezcadena2006] capable of sustaining themselves in a place
+that averages only 200 mm of rainfall every year [@rebman_roberts2012]. This
+circumstance is not helped by the lack of surficial freshwater stocks
+[@maya1997], with the only permanent sources of naturally occurring freshwater
+being arid-land springs [@grismer1993; @macfarlan2020a; @macfarlan_etal2020a;
+@macfarlan2021]. @lerback2022 provides a fine-grained analysis of these spring
+systems in the project area, describing them as primarily lentic pools at low
+points in the arroyos (see @fig-context-photos B).
+
+## Models
+
+To evaluate the determinants of ranch occupation and population size, we rely on
+two generalized linear models (GLMs). We use the first GLM to evaluate whether
+market integration or local ecology has played a larger role in structuring the
+distribution of occupied and abandoned ranch clusters. For this model we assume
+that the dependent variable, whether a ranch cluster is occupied or abandoned,
+has a binomial distribution and is related to the linear predictor by the logit
+link: $$
+\begin{aligned}
+Y &\sim B(n, p) \\
+E[Y/n] &= p \\
+logit(p) &= \beta_{0} + \beta_{1} D(La\; Paz) + \beta_{2} D(Springs)
+\end{aligned}
+$$ with $n$ being the number of ranch clusters, $p$ the probability that a ranch
+cluster is occupied, and $D(x)$ the travel time from a ranch cluster to $x$ (see
+below for more details about how we calculate this).
+
+We use the second GLM to identify which variable contributes more to the
+distribution of rancher populations across currently occupied clusters, with an
+additional variable representing land security, or the proportion of ranch
+houses within a cluster having clear title to their property, as this may
+mediate ranch population size. For this model, we assume that the dependent
+variable, the population size of a ranch cluster, has a negative binomial
+distribution and is related to the linear predictor by the log link: $$
+\begin{aligned}
+Y &\sim NB(\lambda, k) \\
+E[Y] &= \lambda \\
+log(\lambda) &= \beta_{0} + \beta_{1} D(Cities) + \beta_{2}D(Springs) + \beta_{3} P(Security)
+\end{aligned}
+$$ with $\lambda$ being the average number of individuals per ranch cluster, $k$
+a dispersion parameter, and $P(Security)$ the proportion of ranches in a ranch
+cluster that are secure. The use of a negative binomial rather than the standard
+Poisson model for count data is preferred here to account for dispersion, or
+circumstances in which the variance deviates from the mean, which violates a key
+assumption of the Poisson distribution.
+
+Before continuing, we would like to emphasize an important historical dimension
+to this research, namely, that the process leading to the current distribution
+of occupied and abandoned ranch clusters played out before Ciudad Constitucion
+(the large urban area to the north of our study area) offered significant market
+benefits. This is reflected in how we specify each model. The model of occupied
+and abandoned ranch clusters includes only proximity to La Paz, $D(La\; Paz)$,
+as a measure of market integration, which historically was the only market that
+ranchers accessed prior to the construction of Federal Highway 1. The model of
+population distributions across occupied ranches, however, includes travel time
+to La Paz and Ciudad Constitucion, $D(Cities)$, as both have offered markets
+regularly used by ranchers in our study population today.
+
+## Data
+
+The dependent variables, occupation status ($p$) and population size
+($\lambda$), we obtain from Mexico's online Catálogo do Localidades
+[@secretaríadedesarrollosocial2021], coupled with fieldwork conducted over
+several years beginning in 2014. In total, the sample includes 72 ranch
+clusters, with 39 being unoccupied or abandoned and the remaining 33 being
+occupied by 290 individuals. The number of individuals per ranch cluster ranges
+from 1 to 32, with a mean of 8.8. To provide additional context,
+@fig-context-photos B shows an example of an abandoned ranch.
+@fig-context-photos C shows a cluster of currently occupied ranches.
+
+The independent variables, market integration and habitat suitability, we
+operationalize using three estimates of travel time: (i) one to La Paz, (ii) one
+to Ciudad Constitucion or La Paz (depending on the watershed), and (iii) one to
+local springs, with each involving a mix of pedestrian and automotive travel.
+The first two variables are our proxies for market integration, under the
+assumption that increased interaction with markets will positively covary with
+proximity to the cities. The third variable, travel time to springs, is our
+proxy for habitat suitability, as water is a critical resource in this arid
+environment. In the binomial GLM, which examines the distribution of abandoned
+versus active ranching clusters, we use only travel time to La Paz (i) as Ciudad
+Constitucion did not offer a significant market when those locations were being
+established. We then use travel time to both Ciudad Constitucion and La Paz (ii)
+in the negative binomial GLM examining ranch-cluster population sizes. travel
+time to springs (iii) is used in both models, on the assumption that the
+location of active streams would not have changed during the period of time in
+question [@grismer1993].
+
+We define land security as the proportion of ranch houses within a cluster with
+clear title to land or who are registered members of an *ejido*. Information on
+this is provided by Mexico's National Institute of Statistics and Geography
+[@registroagrarionacional2019] and by ethnographic observation. Of the occupied
+ranch clusters, only 8 are fully secure, meaning all ranch houses within the
+cluster have full title to their property. Of the remaining 25, less than half
+of their ranch houses are secure.
+
+A summary of these data can be found in @tbl-variables.
+
+## Travel time
+
+The critical variable in our models is travel time, $D(x)$. This has been used
+as a proxy for market integration in other studies focusing on largely
+subsistence-based economies [see especially @gurven2017], with models exploring
+its relationship to, among other things, social status [@schultz2019], hunting
+effort and efficiency [@luz2015], and oral tradition [@schniter2018]. An
+important advantage of our analysis, however, is that we use travel time as a
+proxy for both market integration and ecosystem services. Specifically, we use
+travel time to major cities as a proxy for market integration and travel time to
+local surficial springs as a proxy for ecosystem services. This allows for more
+straightforward comparisons of their relative contributions to settlement
+outcomes, which can then be readily interpreted as strategies for navigating
+trade-offs between those two variables. The use of these travel-time estimates
+also makes our models analogous to models from urban and economic geography,
+where travel time (or, more generally, transport costs) to and from locations
+providing various goods and services is assumed to be a key determinant of
+residential choice [@alonso1964; @lee_etal2010; @acheampong_silva2015;
+@waddell_etal2007].
+
+To estimate travel time, we first derive a gridded cost surface from a 30-meter
+resolution digital elevation model obtained from the National Elevation Dataset
+provided by the United States Geological Survey. This is achieved by applying
+Campbell's hiking function [@campbell2019] to slope estimates extracted from the
+elevation model. The result is a cost surface where the value of each 30-meter
+grid cell represents the time in seconds it would take to traverse that cell.
+
+Three important assumptions underlie the cost-surface model. First, we assume
+that the average walking speed of a rancher is approximately equal to the third
+decile of US hikers (about 2.5 meters per second over flat ground), which
+increases or decreases based on slope. Second, we assume that no individual will
+traverse a slope greater than 35 degrees. Obviously, there will be exceptions to
+this, but ethnographic observation does suggest that it is true in general for
+the rancheros in this region. Plus, our distance measure is used for relative
+comparisons, making the precise absolute value of limited concern. Finally, we
+assume that travel on roads and built paths will be less costly than through
+grid cells where vegetation and other landform characteristics provide obstacles
+to travel. In the cost surface, this is achieved by first classifying paths as
+primary (for maintained roads), secondary (for non-maintained roads), and
+tertiary (for walking paths), and then weighting the travel cost associated with
+each by 0.2, 0.5, and 0.8, respectively. This is equivalent to increasing the
+speed of travel to 12.5, 5, and 3.125 m/s, making the speeds more comparable to
+driving a car or riding a horse along the primary and secondary paths, which is
+what the ranchers here do when going to the city or visiting their springs or
+neighbors. All road paths were digitized manually using Google Earth and
+validated through field observation.
+
+Once we have this cost surface, we then use Djikstra's algorithm
+[@cormen_etal2001] to identify least-cost paths from ranch clusters to springs
+and cities and, thus, the total time it takes to travel from each origin to each
+destination. These are shown in @fig-shortest-paths. We note, however, that
+strictly speaking, we do not measure travel time to the cities, just to the edge
+of the study region in the direction of Federal Highway 1, for both the northern
+and southern watersheds. Each one has a single road that connects it to Highway
+1 and thus to the markets in Ciudad Constitucion to the north and the capital La
+Paz to the south. The "road" that connects the watersheds to each other is
+poorly maintained, difficult to traverse, and only used out of necessity. As a
+consequence, nearly all of the variance in travel time to cities is found within
+each watershed. For that reason, we estimate the time it takes to leave each
+watershed ("leaving" here means hitting the edge of the project area), rather
+than the true time it takes to get to each city. We interpret this as the
+relative difference in time it takes to get to Ciudad Constitucion from the
+northern watershed and the relative difference in time it takes to get to La Paz
+from the southern watershed. To get to La Paz from the northern watershed, we
+simply add an arbitrary half hour to the total time to get out of that
+watershed. The same would presumably be the case to get to Ciudad Constitucion
+from the southern watershed, though that value does not factor into our
+analysis. For springs, our ethnographic data reveal that individuals residing
+within the same ranch cluster tend to rely on one or two springs at most
+[@lerback2022]. Thus, we estimate road distance to the two nearest springs and
+take the average.
+
+## Model evaluation
+
+For completeness, we fit three versions of the negative binomial model of ranch
+cluster population: one without the land security variable $P(Secure)$, one with
+it included as an additive term, and one with it included as an interaction
+term. After fitting these models, we perform an ANOVA Likelihood Ratio Test for
+significant improvement in fit when adding each version of the land security
+variable. We then test for spatial autocorrelation in the untransformed
+residuals of all models, including the binomial model, using Monte Carlo
+simulations of Moran's I. We also test for multicollinearity in our variables by
+calculating their respective variance inflation factors.
+
+All analyses are conducted in the R programming language and environment
+[@rcoreteam2023]. For more details on these models and the travel time analysis,
+please see the supplementary materials.
+
+# Results
+
+We initiated this research to understand how local ecology and market
+integration influenced settlement dynamics. Our first model examines how these
+two factors influence whether or not a ranch cluster is occupied or abandoned.
+The model suggests that a cluster's status (whether it is occupied or abandoned)
+negatively co-varies with travel time to spring ($\beta$ = -4.803, *p* = 0.009)
+and positively co-varies with travel time to La Paz ($\beta$ = 2.797, *p* \<
+0.001), indicating that ranch clusters are more likely to be occupied closer to
+springs and further from town (@tbl-coefficients, @fig-response-plots). Our
+second model examines how local ecology and market integration influence the
+distribution of the population across occupied ranch clusters today. The model
+shows that population size across occupied ranch clusters negatively co-varies
+with travel time to cities ($\beta$ = -0.802, *p* \< 0.001), for both
+Constitucion and La Paz, but does not significantly co-vary with travel time to
+springs ($\beta$ = 0.699, *p* = 0.299) (@tbl-coefficients, @fig-response-plots).
+The model with an additive land security term suggests that more secure ranch
+clusters have slightly smaller populations, but the effect is non-significant
+($\beta$ = -0.114, *p* = 0.687). The ANOVA also indicates that an additive land
+security term does not significantly improve the model ($\chi$(1) = 3.003, *p* =
+0.684). The same is true for an interactive term ($\chi$(1) = 3.130, *p* =
+0.663).
+
+The null hypothesis of no spatial autocorrelation in the untransformed residuals
+cannot be rejected for either the occupation status (*I* = -0.023, *p* = 0.820)
+or population size (*I* = -0.082, *p* = 0.460) model, thus indicating a lack of
+spatial correlation. Measures of variance inflation are minimal in each case,
+approximately one for all terms, strongly suggesting no multicollinearity in the
+independent variables.
+
+For a complete list of summary statistics for each model, see @tbl-models.
+
+# Discussion
+
+With this analysis, we seek to understand how rural ranchers in BCS respond to
+important trade-offs between their local ecology and market integration. Our
+results indicate that the locations of modern ranch clusters are largely
+constrained by access to surficial water sources, our proxy for habitat
+suitability in this arid environment, rather than access to the capital La Paz.
+This observation is supported by the fact that occupied ranch clusters are more
+likely to be found farther from the capital, rather than closer to it. However,
+within those constraints, the rancheros are now living at larger densities in
+locations with easier access to Ciudad Constitucion and La Paz, our proxies for
+market integration. In other words, water availability determines *where* people
+live but not *how many* people live there. Instead, that is being driven largely
+by markets, which pull people closer to the cities.
+
+## Proxies for market integration
+
+One important advantage of this research concerns the use of travel time as a
+measure of both market integration and local adaptation. Using this simple
+metric has a number of advantages. First, it provides a straightforward way of
+evaluating the relative contribution of each variable to the outcome of
+interest, in this case, settlement decisions among rural ranchers. Second, it
+conforms with geographic theory regarding spatial interactions and dependence
+(otherwise known as Tobler's Law, @tobler1970) and overcomes known limitations
+of Euclidean distance as a cost measure. Finally, it is a relatively simple
+matter to derive those cost estimates from easily accessible remote-sensed
+elevation and road network data held in online repositories. High resolution
+informant data, on the other hand, are generally difficult and costly to
+acquire, often with little chance of providing the right sort of overlap between
+different variables required for fair comparisons.
+
+It is true, as @mattison2022 point out, that market integration is a
+multi-dimensional process, making it at first blush an awkward concept to try
+and capture in a single variable like travel time or caloric-intake
+[@henrich2010]. Still, Tobler's Law would seem to suggest that all the variables
+strongly associated with market integration will also co-vary with proximity to
+cities and their markets. In other words, the costs and benefits of market
+integration will positively co-vary with access to markets - both will increase
+as one gains greater access to and interacts more often with markets. So, the
+closer individuals are, the more likely they should be to derive their caloric
+requirements from cities, to work in cities, to own televisions, cellphones, and
+other goods sold in cities, but also the more likely they should be to face
+intense competition for jobs, higher rents, and other costs associated with
+living in that proximity.
+
+The ranchers in our study population also recognize that those who live closer
+to the trans-peninsular highway have greater access to markets and are more
+likely to interact with them. They have said as much during our conversations
+with them. As mentioned above, travel time is also a fundamental variable in
+urban and economic geography [@krugman1991; @fujita2001] and thus finds support
+from that direction, too. So, when we look at where individuals actually choose
+to live (how near or how far from cities and markets they choose to reside), it
+is safe to assume we are also getting an answer to the question of how they
+weigh the costs and benefits of cities against those afforded by a rural way of
+life.
+
+## Historical antecedents
+
+What explains our results? There are many dimensions to this question, but key
+to any general understanding of such processes is the idea that people will
+choose to live where they can make the best living for themselves given the
+socioeconomic and environmental constraints that confront them. Those
+constraints will, in turn, differ depending on how one chooses to go about
+making a living [@vernon2022]. For the people of the Sierra de la Giganta, that
+means living where you can expect to do ranching and basic crop production
+reasonably well, that is, where one has access to permanent water sources, as
+well as flat lands that sit above dry riverbeds (so they are not impacted by
+flooding caused by seasonal tropical storms and hurricanes). Due to the lack of
+meaningful water infrastructure in this region, households depend on these
+springs to support both domestic (e.g. bathing and drinking) and economic needs
+(e.g. supporting livestock and crop production). Our statistical results are
+consistent with this idea and with our ethnographic interviews with ranchers who
+indicate that access to a spring is the single most important factor for
+establishing and maintaining a home [@lerback2022].
+
+Our first statistical model indicates that abandoned ranches were located
+farther from springs. Although this appears to run counter to our previous claim
+about the importance of springs to all settlement in the region, we believe this
+is an artifact of our analytic framework, which required us to collapse the
+temporal dimension of this largely historical process. Historical and
+ethnographic evidence suggest that as the first Euro-American colonists entered
+this region in the 18th century, they occupied the largest springs in close
+proximity to arable land [@crosby1994; @macfarlan_etal2020;
+@macfarlan_etal2021]. As the population expanded over time, individuals would
+move up or down drainages, to the next closest spring that could support
+subsistence needs. After time, ranchers likely saturated the landscape by
+occupying all areas with permanent freshwater [sensu @crosby1981] as well as
+less desirable locations which were supported not by springs, but by ephemeral
+freshwater sources such as tinajas and seasonally filled lake basins, which can
+hold water for months or years if they receive sufficient precipitation via the
+North American Monsoon.
+
+However, after the construction of the Transpeninsular Highway and associated
+graded dirt roads in the 1970s and 1980s, the payoffs structuring settlement
+decisions were likely altered, as novel constraints and opportunities emerged
+[@barkenbus1974; @mathes2015]. For a smallholder population undergoing market
+transition, the ideal household location has become both next to a spring and
+close to a road. These roads allowed ranchers to access urban goods and services
+more easily and created an incentive for households occupying springs and
+ephemeral freshwater sources far from the road to move closer to it
+[@macfarlan_etal2023a]. This process resulted in a depopulation event between
+the two major drainages. As this process has unfolded over the last 30-40 years,
+springs near roads with the closest access to urban markets have experienced
+increasing intensification as suitable habitats for establishing new households
+have been exhausted, resulting in the highest population densities.
+
+## Land Security
+
+Our measure of land security proved to be non-significant in our model. Rather
+than being an outcome from which to draw conclusions, we believe this result
+should be interpreted with caution. Our measure of land security is based on a
+combination of government data and informant observations and so likely does not
+capture the full complexity of this important issue. A larger dataset with a
+more standardized and quantifiable measure of land security would be more
+representative. We also note that the issue of land security is not equally
+salient to everyone in this population. Many of the informants that might be
+land insecure by most objective measures do not necessarily act as if their
+title to land is uncertain. This would suggest that perceptions of risk with
+respect to property rights should be studied more thoroughly.
+
+## Socio-ecological consequences
+
+The findings presented here have important consequences for traditional ranching
+communities living in similar circumstances, for how they maintain their way of
+life while meeting the challenges posed by climate change, land fragmentation,
+and other exogenous forces. For instance, researchers have shown that market
+integration within traditional agricultural communities involves a shift from
+intensive to extensive kinship systems, or shifts from small kin networks with
+higher degrees of relatedness to large kin networks with lower degrees of
+relatedness [@peña2002]. This appears to be the result of a decreasing reliance
+on consanguineous or close-kin marriage customs, with families favoring
+expansive social networks within a market economy over stronger family ties
+within a largely agrarian economy [@shenk2016]. Currently, the ranching
+communities of BCS are characterized by more intensive kin networks, but the
+limited available evidence suggests that this might be changing, with
+individuals increasingly marrying outside of the ranching community, most likely
+to individuals living within the cities. Families are as a consequence
+increasing their connections to the cities and by extension increasing the
+incentives for integrating more closely with their markets.
+
+These settlement dynamics have also had variable ecosystem outcomes. Springs
+that exist closer to the road, which are more heavily utilized, typically
+experience substantial negative outcomes relative to springs that exist in the
+depopulated zone, including biodiversity loss, higher rates of introduced exotic
+species, soil degradation, and noise and chemical pollution
+[@leondelaluz_dominguezcadena2006]. This is unfortunate, as the springs and the
+emergent terrestrial ecosystems they support are considered Ramsar sites
+(wetlands of international importance), as well as national terrestrial and
+hydrological priority regions [@arriagacabrera2002; @ramsar2022]. However, it is
+inaccurate to assume, as many conservation biologists, NGOs, and urban elites
+seem to do, that ranchers and their lifestyles have uniformly negative impacts
+on regional ecosystem health [@riemann2005]. For it appears that small, riparian
+ecosystems surrounding springs in the depopulated hinterland away from roads are
+currently on the rebound, with higher rates of endemism, and this in spite of
+the fact that ranching continues in the region. Similar outcomes have also been
+noted elsewhere [@vaccaro2021]. So, concentration of the population within this
+region might actually be beneficial, at least from the perspective of
+biodiversity and ecosystem health, a pattern that appears to be recurring across
+the globe as populations become more urbanized [@daskalova2023]. For that
+matter, current research suggests that the complete abandonment of a region can
+have negative consequences for biodiversity given the co-evolutionary histories
+of small-scale subsistence economies and their local ecosystems [@fischer2012].
+
+Current demographic trends across the region may also exacerbate these local
+processes [@vasco2017]. Since the signing of NAFTA, the total population of the
+southern peninsula has more than doubled, from 375 thousand in 1995 to nearly
+800 thousand in 2020. As one might expect, these increases have occurred almost
+exclusively within the peninsula's urban corridors [@inegi2023a; @inegi2023] .
+This entails increased demand for goods and services and by necessity, increased
+demand for the labor that can provide them, thereby increasing the incentives
+for locating in closer proximity to cities and abandoning more remote locations.
+As alluded to above, the ecological consequences will be complex and
+multifaceted. Given the complexities of population decline and its consequences
+for biodiversity, it is hard to say whether these demographic trends will have a
+net positive or negative effect on long term ecosystem health. We hope our
+analyses here catalyze future research on the effects of ranch abandonment on
+ecosystem outcomes.
+
+## Climate change and drought
+
+Of particular urgency for the rural ranchers of the Sierra de la Giganta is
+ongoing climate-induced drought [@cavazos2012; @colorado-ruiz2018]. For the
+reasons we have emphasized throughout, a lack of access to government services
+and infrastructure makes them uniquely sensitive to these local impacts
+[@hansen2019]. That is almost certainly why access to water plays a larger role
+in constraining the location of occupied ranch clusters. It may also explain the
+decisions ranchers make about the size and composition of their herds, as intra-
+and inter-annual variation in precipitation - in the absence of water
+infrastructure - is the primary factor driving the abundance and distribution of
+most animal fodder in the region [@pettorelli2013; @macfarlan_etal2023a].
+Previous research has shown that pastoralists respond to and oftentimes prepare
+for drought conditions by reducing their herd size, selling off stock in the
+nearby cities, or switching to more drought tolerant alternatives like small
+ruminants [@mace1989; @mace1990; @mace1993]. While these animals are more
+efficient in terms of their total land area, feed, and water requirements, there
+is less demand for them in city markets. This implies that ranchers in the
+region, while trying to address the ongoing threat of droughts, will lose
+important sources of capital they would need to access other resources provided
+by modern markets that they depend on.
+
+## Conclusion
+
+Trade-offs between local ecology and market integration will continue to shape
+the development of rural, smallholder communities, especially as globalization
+and urbanization accelerate. This is as true of the rural ranchers of the Sierra
+de la Giganta as it is for ranchers, pastoralists, and small-scale
+agriculturalists in other parts of the world. For ranchers living in the remote
+corners of BCS, the data coupled with informant reporting together suggest that
+these ranchers feel the pull of markets, being compelled to infill around
+ranches nearer to roads that give them greater access to those markets. The
+consequences of this spatial sorting of the population are complicated as well.
+While integration can help to mitigate local risks, it also raises new ones,
+with population packing increasing pressure on nearby water resources that are
+crucial to the sustainability of these communities. At the same time, the
+concentration of the population appears to offer positive results for the
+ecology of abandoned areas, among other things increasing their biodiversity.
+
+The research presented in this paper raises a number of important issues that
+deserve greater attention, including a more robust analysis of land
+fragmentation and land insecurity, using a larger dataset more representative of
+these ranching communities across the peninsula. It is also important that we
+better understand the manner and extent to which proximity or travel time to
+markets correlates with other proxies for market integration. Answering that
+question would go along way towards helping researchers overcome many of the
+methodological challenges associated with richly ethnographic studies of remote,
+smallholder economies and in the process offer new avenues for investigating the
+ongoing integration of those systems with modern markets.
+
+{{< pagebreak >}}
+
+# Funding and/or Competing interests
+
+The research was funded in part by the Center for Latin American Studies,
+University of Utah.
+
+The authors have no relevant financial or non-financial interests to disclose.
+
+{{< pagebreak >}}
+
+# References
+
+::: {#refs}
+:::
+
+{{< pagebreak >}}
 
@@
 here("figures", "overview-map.png") |> include_graphics()
-```
-
-Importantly, we investigate the degree to which modern land fragmentation, specifically the security of land title, serves as an effective constraint on these dynamics.
-This is important because pastoralists have historically lacked clear title to land, instead relying on open rangeland systems [@behnke1994; @turner1999].
-In Mexico, these systems take the form of *ejidos*, or areas of communal agricultural land.
-The first *ejidos* were established by the Mexican state in the early part of the twentieth century [@perramond2008], though in our project area they did not arrive until 1975 [@dof1975] and 1976 [@dof1976], respectively.
-While these *ejidos* have historically served as a buffer against land fragmentation, allowing rancheros to maintain their way of life, their ability to play that role continues to change as the pace of privatization and shifting land tenure accelerates [@lesorogol2003; @reid2008].
-As has happened in other contexts, privatization efforts are often led by NGOs, commercial interests, and other external forces, with the result being greater fragmentation of pastoral rangeland systems [@burnsilver2008; @galvin2009].
-For the rural ranchers of the Sierra de la Giganta, this process of fragmentation at least partially explains why ranches of related individuals tend to cluster.
-Rancheros, unable to find new land for themselves, simply build ranches next to their kin on land that is at least historically associated with their family.
-
-Our analysis relies on two generalized linear models (GLMs).
-We use a binomial GLM to evaluate whether market integration or local ecology has played a larger role in determining the current status of ranch clusters, if they are occupied or abandoned.
-We then use a negative binomial GLM to identify which variable contributes more to the distribution of these ranching populations across currently occupied clusters, with a binary term representing land security.
-These are applied to ranch clusters in two east-west running watersheds along the southern extent of the Sierra de la Giganta range.
-
-The key measure in all these models is cost-distance or travel time.
-This has been used as a proxy for market integration in other studies focusing on largely subsistence-based economies [see especially @gurven2017], with models exploring its relationship to, among other things, social status [@schultz2019], hunting effort and efficiency [@luz2015], and oral tradition [@schniter2018].
-An important advantage of our analysis, however, is that we use cost-distance as a proxy for both market integration and local ecology (or ecosystem services).
-Specifically, we use cost-distance to major cities as a proxy for market integration and cost-distance to local surficial springs as a proxy for ecosystem services.
-This allows for more straightforward comparisons of their relative contributions to settlement outcomes, which can then be readily interpreted as strategies for navigating trade-offs between those two variables.
-
-We would also like to emphasize an important historical dimension to this research, namely, that the process leading to the current distribution of occupied and abandoned ranch clusters played out before Ciudad Constitucion (the large urban area to the north of our study area) offered significant market benefits.
-This is reflected in how we specify each model.
-The model of occupied and abandoned ranch clusters includes only proximity to La Paz as a measure of market integration, which historically was the only market that ranchers accessed prior to the construction of Federal Highway 1.
-The model of population distributions across occupied ranches, however, includes cost-distance to La Paz and Ciudad Constitucion, as both have offered markets regularly used by ranchers in our study population today.
-All models are evaluated for goodness of fit, multicollinearity, and residual autocorrelation.
-
-# Results
-
-We initiated this research to understand how local ecology and market integration influenced settlement dynamics.
-Our first model examines how these two factors influence whether or not a ranch cluster is occupied or abandoned.
-The model suggests that a cluster's status (whether it is occupied or abandoned) negatively co-varies with cost-distance to spring ($\beta$ = -4.803, *p* = 0.009) and positively co-varies with cost-distance to La Paz ($\beta$ = 2.797, *p* \< 0.001), indicating that ranch clusters are more likely to be occupied closer to springs and further from town (@tbl-models, @fig-response-plots).
-Our second model examines how local ecology and market integration influence the distribution of the population across occupied ranch clusters today.
-The model shows that population size across occupied ranch clusters negatively co-varies with cost-distance to cities ($\beta$ = -0.802, *p* \< 0.001), for both Constitucion and La Paz, but does not significantly co-vary with cost-distance to springs ($\beta$ = 0.699, *p* = 0.299) (@tbl-models, @fig-response-plots).
-The model with an additive land security term suggests that more secure ranch clusters have slightly smaller populations, but the effect is non-significant ($\beta$ = -0.114, *p* = 0.687).
-The ANOVA also indicates that an additive land security term does not significantly improve the model ($\chi$(1) = 3.003, *p* = 0.684).
-The same is true for an interactive term.
-
-The null hypothesis of no spatial autocorrelation in the untransformed residuals cannot be rejected for either the occupation status (*I* = -0.023, *p* = 0.820) or population size (*I* = -0.082, *p* = 0.460) model, thus indicating a lack of spatial correlation.
-Measures of variance inflation are minimal in each case, approximately one for all terms, strongly suggesting no multicollinearity in the independent variables.
-
-```{r}
-#| label: tbl-models
-
-here("data", "model-statistics.rdata") |> load()
-
-coefficient_estimates |>
-  filter(model %in% c("Occupied-Abandoned", "Population")) |>
-  group_by(model) |> 
-  gt() |> 
-  tab_options(
-    column_labels.font.weight = "bold",
-    data_row.padding = px(6),
-    heading.align = "left",
-    heading.border.bottom.style = "hidden",
-    row_group.as_column = TRUE,
-    row.striping.background_color = "gray96",
-    row.striping.include_stub = FALSE,
-    row.striping.include_table_body = TRUE,
-    table.border.top.style = "hidden",
-    table.border.bottom.style = "hidden"
-  ) |> 
-  sub_missing(missing_text = "") |> 
-  tab_stubhead("model") |> 
-  tab_header(title = "Coefficient Estimates") |> 
-  tab_caption("") |> 
-  as_latex()
 
@@
 
-# Discussion
+{{< pagebreak >}}
 
-With this analysis, we seek to understand how rural ranchers in Baja California Sur respond to important trade-offs between their local ecology and market integration.
-Our results indicate that the locations of modern ranch clusters are largely constrained by access to surficial water sources, our proxy for habitat suitability in this arid environment, rather than access to the capital La Paz.
-This observation is supported by the fact that occupied ranch clusters are more likely to be found farther from the capital, rather than closer to it.
-However, within those constraints, the rancheros are now living at larger densities in locations with easier access to Ciudad Constitucion and La Paz, our proxies for market integration.
-In other words, water availability determines *where* people live but not *how many* people live there.
-Instead, that is being driven largely by markets, which pull people closer to the cities.
+```{r}
+#| label: fig-shortest-paths 
+#| fig-align: center 
+#| out-width: 5.5in 
+#| fig-cap: "**Shortest paths.** Maps showing least-cost paths to springs and cities."
+
+here("figures", "least-cost-paths.png") |> include_graphics()
+
+```
+
+{{< pagebreak >}}
 
@@
 #| fig-align: center
-#| out-width: 5in
+#| out-width: 5.5in
 #| fig-cap: "**Response plots.** Marginal response plots showing change in the dependent variables (occupied vs abandoned ranch clusters and population size) as a function of change in one independent variable while holding the other at its mean."
@@
 here("figures", "response-plots.png") |> include_graphics()
+
 ```
-
-## Proxies for market integration
-
-One important advantage of this research concerns the use of travel costs as a measure of both market integration and local adaptation.
-Using this simple metric provides a straightforward way of evaluating the relative contribution of each variable to the outcome of interest, in this case, settlement decisions among rural ranchers.
-It also conforms with geographic theory regarding spatial interactions and dependence (otherwise known as Tobler's Law) and overcomes known limitations of Euclidean distance as a cost measure.
-And, it is a simple matter to derive those cost estimates from easily accessible remote-sensed elevation and road network data held in online repositories.
-High resolution informant data, on the other hand, are generally difficult and costly to acquire, often with little chance of providing the right sort of overlap between different variables required for fair comparisons.
-
-It is true, as @mattison2022 point out, that market integration is a multi-dimensional process, making it at first blush an awkward concept to try and capture in a single variable like cost-distance or caloric-intake [@henrich2010].
-Still, Tobler's Law would seem to suggest that all the variables strongly associated with market integration will also co-vary with proximity to cities and their markets.
-In other words, the costs and benefits of market integration will positively co-vary with access to markets - both will increase as one gains greater access to and interacts more often with markets.
-So, the closer individuals are, the more likely they should be to derive their caloric requirements from cities, to work in cities, to own televisions, cellphones, and other goods sold in cities, but also the more likely they should be to face intense competition for jobs, higher rents, and other costs associated with living in that proximity.
-
-The ranchers in our study population also recognize that those who live closer to the trans-peninsular highway have greater access to markets and are more likely to interact with them.
-They have said as much during our conversations with them.
-It is perhaps worth noting in this context that cost distance (specifically transport cost) is also a fundamental variable in the new economic geography, where models seek to explain urbanization or the spatial concentration of economic activity [@krugman1991; @fujita2001] and thus finds support from that direction, too.
-So, when we look at where individuals actually choose to live (how near or how far from cities and markets they choose to reside), it is safe to assume we are also getting an answer to the question of how they weigh the costs and benefits of cities against those afforded by a rural way of life.
-
-## Historical antecedents
-
-What explains our results?
-There are many dimensions to this question, but key to any general understanding of such processes is the idea that people will choose to live where they can make the best living for themselves given the socioeconomic and environmental constraints that confront them.
-Those constraints will, in turn, differ depending on how one chooses to go about making a living [@vernon2022].
-For the people of the Sierra de la Giganta, that means living where you can expect to do ranching and subsistence crop production reasonably well, that is, where one has access to permanent water sources, as well as flat lands that sit above dry riverbeds (so they are not impacted by flooding caused by seasonal tropical storms and hurricanes).
-Much like the vast majority of the Baja California peninsula, the Sierra de La Giganta is a hot, arid, mountainous landscape with minimal surficial freshwater stocks [@maya1997].
-The only permanent sources of naturally occurring freshwater in this region are arid-land springs [@grismer1993; @macfarlan2020a; @macfarlan2020b; @macfarlan2021].
-Due to the lack of infrastructure in this region, households require springs to support both domestic (e.g.
-bathing and drinking) and economic needs (e.g.
-supporting livestock and crop production).
-Our statistical results are consistent with our ethnographic interviews with ranchers which indicate that access to a spring is the single most important factor for establishing and maintaining a home [@lerback2022].
-
-Our first statistical model indicates that abandoned ranches were located farther from springs.
-Although this appears to run counter to our previous claim about the importance of springs to all settlement in the region, we believe this is an artifact of our analytic framework, which required us to collapse the temporal dimension of this largely historical process.
-Historical and ethnographic evidence suggest that as the first Euro-American colonists entered this region in the 18th century, they occupied the largest springs in close proximity to arable land [@crosby1994].
-As the population expanded over time, individuals would move up or down drainages, to the next closest spring that could support subsistence needs.
-After time, ranchers likely saturated the landscape by occupying all areas with permanent freshwater [sensu @crosby1981] as well as less desirable locations which were supported not by springs, but by ephemeral freshwater sources such as tinajas and seasonally filled lake basins, which can hold water for months or years if they receive sufficient precipitation via the North American Monsoon.
-
-However, after the construction of the Transpeninsular Highway and associated graded dirt roads in the 1970s and 1980s, the payoffs structuring settlement decisions were likely altered, as novel constraints and opportunities emerged.
-For a subsistence population undergoing market transition, the ideal household location has become both next to a spring and close to a road.
-These roads allowed ranchers to access urban goods and services more easily and created an incentive for households occupying springs and ephemeral freshwater sources far from the road to move closer to it.
-This process resulted in a depopulation event between the two major drainages.
-As this process has unfolded over the last 30-40 years, springs near roads with the closest access to urban markets have experienced increasing intensification as suitable habitats for establishing new households have been exhausted, resulting in the highest population densities.
-
-## Socio-ecological consequences
-
-The findings presented here have important consequences for traditional ranching communities living in similar circumstances, for how they maintain their way of life while meeting the challenges posed by climate change, land fragmentation, and other exogenous forces.
-For instance, researchers have shown that market integration within traditional agricultural communities involves a shift from intensive to extensive kinship systems, or shifts from small kin networks with higher degrees of relatedness to large kin networks with lower degrees of relatedness [@peña2002].
-This appears to be the result of a decreasing reliance on consanguineous or close-kin marriage customs, with families favoring expansive social networks over stronger family ties within a market economy [@shenk2016].
-Currently, the ranching communities of Baja California Sur are characterized by more intensive kin networks, but the limited available evidence suggests that this might be changing, with individuals increasingly marrying outside of the ranching community, most likely to individuals living within the cities.
-Families are as a consequence increasing their connections to the cities and by extension increasing the incentives for integrating more closely with their markets.
-
-These settlement dynamics have also had variable ecosystem outcomes.
-Springs that exist closer to the road, which are more heavily utilized, typically experience substantial negative outcomes relative to springs that exist in the depopulated zone, including biodiversity loss, higher rates of introduced exotic species, soil degradation, and noise and chemical pollution [@leóndelaluz2006].
-This is unfortunate, as the springs and the emergent terrestrial ecosystems they support are considered Ramsar sites (wetlands of international importance), as well as national terrestrial and hydrological priority regions [@arriagacabrera2002; @ramsar2022].
-However, it is inaccurate to assume, as many conservation biologists, NGOs, and urban elites seem to do, that ranchers and their lifestyles have uniformly negative impacts on regional ecosystem health [@riemann2005].
-For it appears that small, riparian ecosystems surrounding springs in the depopulated hinterland away from roads are currently on the rebound, with higher rates of endemism, and this in spite of the fact that ranching continues in the region.
-Similar outcomes have also been noted elsewhere [@vaccaro2021].
-So, concentration of the population within this region might actually be beneficial, at least from the perspective of biodiversity and ecosystem health, a pattern that appears to be recurring across the globe as populations become more urbanized [@daskalova2023].
-For that matter, current research suggests that the complete abandonment of a region can have negative consequences for biodiversity given the co-evolutionary histories of small-scale subsistence economies and their local ecosystems [@fischer2012].
-
-Current demographic trends across the region may also exacerbate these local processes [@vasco2017].
-Since the signing of NAFTA, the total population of the southern peninsula has more than doubled, from 375 thousand in 1995 to nearly 800 thousand in 2020.
-As one might expect, these increases have occurred almost exclusively within the peninsula's urban corridors [@inegi2023a; @inegi2023] .
-This entails increased demand for goods and services and by necessity, increased demand for the labor that can provide them, thereby increasing the incentives for locating in closer proximity to cities and abandoning more remote locations.
-
-## Climate change and drought
-
-Of particular urgency for the rural ranchers of the Sierra de la Giganta is ongoing climate-induced drought [@cavazos2012; @colorado-ruiz2018].
-For the reasons we have emphasized throughout, a lack of access to government services and infrastructure makes them uniquely sensitive to these local impacts [@hansen2019].
-That is almost certainly why access to water plays a larger role in constraining the location of occupied ranch clusters.
-It may also explain the decisions ranchers make about the size and composition of their herds, as intra- and inter-annual variation in precipitation - in the absence of water infrastructure - is the primary factor driving the abundance and distribution of most animal fodder in the region [@pettorelli2013; @macfarlan2023].
-Previous research has shown that pastoralists respond to and oftentimes prepare for drought conditions by reducing their herd size, selling off stock in the nearby cities, or switching to more drought tolerant alternatives like small ruminants [@mace1989; @mace1990; @mace1993].
-While these animals are more efficient in terms of their total land area, feed, and water requirements, there is less demand for them in city markets.
-This implies that ranchers in the region, while trying to address the ongoing threat of droughts, will lose important sources of capital they would need to access other resources provided by modern markets that they depend on.
-
-## Conclusion
-
-Trade-offs between local ecology and market integration will continue to shape the development of rural, largely subsistence-based communities, especially as globalization and urbanization accelerate.
-This is as true of the rural ranchers of the Sierra de la Gigantes as it is for ranchers, pastoralists, and small-scale agriculturalists in other parts of the world.
-For ranchers living in the remote corners of Baja California Sur, the data coupled with informant reporting together suggest that these ranchers feel the pull of markets, being compelled to infill around ranches nearer to roads that give them greater access to those markets.
-The consequences of this spatial sorting of the population are complicated as well.
-While integration can help to mitigate local risks, it also raises new ones, with population packing increasing pressure on nearby water resources that are crucial to the sustainability of these communities.
-At the same time, the concentration of the population appears to offer positive results for the ecology of abandoned areas, among other things increasing their biodiversity.
-
-The research presented in this paper raises a number of important issues that deserve greater attention, particularly the manner and extent to which proximity or cost-distance to markets correlates with other proxies for integration.
-Answering that question would go along way towards helping researchers overcome many of the methodological challenges associated with richly ethnographic studies of remote and largely subsistence-based economies and in the process offer new avenues for investigating the ongoing integration of those systems with modern markets.
-
-# Methods
-
-To evaluate the determinants of ranch occupation and population size, we rely on two generalized linear models (GLMs).
-We use a binomial GLM to evaluate whether market integration or local ecology has played a larger role in structuring the distribution of occupied and abandoned ranch clusters.
-We then use a negative binomial GLM to identify which variable contributes more to the distribution of rancher populations across currently occupied clusters, with an additional variable representing land security, or the proportion of ranch houses within a cluster having clear title to their property, as this may mediate ranch population size.
-The use of a negative binomial rather than the standard Poisson model for count data is preferred here to account for dispersion, or circumstances in which the variance deviates from the mean, which violates a key assumption of the Poisson distribution.
-These models are applied to ranch clusters in two largely east-west running watersheds along the southern extent of the Sierra de la Giganta range.
-
-## Data
-
-The dependent variables, occupation status and population, we obtain from Mexico's online Catálogo do Localidades [@secretaríadedesarrollosocial2021], coupled with fieldwork conducted over several years beginning in 2014.
-In total, the sample includes 72 ranch clusters, with 39 being unoccupied or abandoned and the remaining 33 being occupied by 290 individuals.
-
-The independent variables, market integration and habitat suitability, we operationalize using three estimates of cost-distance: (i) one to La Paz, (ii) one to Ciudad Constitucion or La Paz (depending on the watershed), and (iii) one to local springs, with each involving a mix of pedestrian and automotive travel.
-The first two variables are our proxies for market integration, under the assumption that increased interaction with markets will positively covary with proximity to the cities.
-The third variable, cost-distance to springs, is our proxy for habitat suitability, as water is a critical resource in this arid environment.
-In the binomial GLM, which examines the distribution of abandoned versus active ranching clusters, we use only cost-distance to La Paz (i) as Ciudad Constitucion did not offer a significant market when those locations were being established.
-We then use cost-distance to both Ciudad Constitucion and La Paz (ii) in the negative binomial GLM examining ranch-cluster population sizes.
-Cost-distance to springs (iii) is used in both models, on the assumption that the location of active streams would not have changed during the period of time in question [@grismer1993].
-
-Land fragmentation is here operationalized as land security, defined as the proportion of ranch houses within a cluster with clear title to land or who are registered members of an ejido.
-Information on this is provided by Mexico's National Institute of Statistics and Geography [@registroagrarionacional2019] and by ethnographic observation.
-Of the occupied ranch clusters, only 8 are fully secure, meaning all ranch houses within the cluster have full title to their property.
-Of the remaining 25, less than half of their ranch houses are secure.
-
-## Travel time
-
-To estimate travel time, we first derive a gridded cost surface from a 30-meter resolution digital elevation model derived from the National Elevation Dataset provided by the United States Geological Survey.
-This is achieved by applying Campbell's hiking function [@campbell2019] to slope estimates extracted from the elevation model.
-The result is a cost surface where the value of each 30-meter grid cell represents the time in seconds it would take to traverse that cell.
-
-Three important assumptions underlie the cost-surface model.
-First, we assume that the average walking speed of a rancher is approximately equal to the third decile of US hikers (about 2.5 meters per second over flat ground), which increases or decreases based on slope.
-Second, we assume that no individual will traverse a slope greater than 35 degrees.
-Obviously, there will be exceptions to this, but ethnographic observation does suggest that it is true in general for the rancheros in this region.
-Plus, our distance measure is used for relative comparisons, making the precise absolute value of limited concern.
-Finally, we assume that travel on roads and built paths will be less costly than through grid cells where vegetation and other landform characteristics provide obstacles to travel.
-In the cost surface, this is achieved by first classifying paths as primary (for maintained roads), secondary (for non-maintained roads), and tertiary (for walking paths), and then weighting the travel cost associated with each by 0.2, 0.5, and 0.8, respectively.
-This is equivalent to increasing the speed of travel to 12.5, 5, and 3.125 m/s, making the speeds more comparable to driving a car or riding a horse along the primary and secondary paths, which is what the ranchers here do when going to the city or visiting their springs or neighbors.
-All road paths were digitized manually using Google Earth and validated through field observation.
-
-```{r}
-#| label: fig-shortest-paths 
-#| fig-align: center 
-#| out-width: 5.75in 
-#| fig-cap: "**Shortest paths.** Maps showing least-cost paths to springs and cities."
-
-here("figures", "least-cost-paths.png") |> include_graphics()
-```
-
-Once we have this cost surface, we then use Djikstra's algorithm to identify least-cost paths from ranch clusters to springs and cities and, thus, the total time it takes to travel from each origin to each destination.
-These are shown in @fig-shortest-paths.
-We note, however, that strictly speaking, we do not measure travel time to the cities, just to the edge of the study region in the direction of Federal Highway 1, for both the northern and southern watersheds.
-Each one has a single road that connects it to Highway 1 and thus to the markets in Ciudad Constitucion to the north and the capital La Paz to the south.
-The "road" that connects the watersheds to each other is not maintained and thus rarely used.
-As a consequence, nearly all of the variance in travel time to cities is found within each watershed.
-For that reason, we estimate the time it takes to leave each watershed ("leaving" here means hitting the edge of the project area), rather than the true time it takes to get to each city.
-We interpret this as the relative difference in time it takes to get to Ciudad Constitucion from the northern watershed and the relative difference in time it takes to get to La Paz from the southern watershed.
-To get to La Paz from the northern watershed, we simply add an arbitrary half hour to the total time to get out of that watershed.
-The same would presumably be the case to get to Ciudad Constitucion from the southern watershed, though that value does not factor into our analysis.
-For springs, our ethnographic data reveal that individuals residing within the same ranch cluster tend to rely on one or two springs at most [@lerback2022].
-Thus, we estimate road distance to the two nearest springs and take the average.
-
-## Model evaluation
-
-For completeness, we fit three versions of the negative binomial model of ranch cluster population: one without the land security variable, one with it included as an additive term, and one with it included as an interaction term.
-After fitting these models, we perform an ANOVA Likelihood Ratio Test for significant improvement in fit.
-We then test for spatial autocorrelation in the untransformed residuals of all models, including the binomial model, using Monte Carlo simulations of Moran's I.
-We also test for multicollinearity in our cost-distance variables by calculating their respective variance inflation factors.
-
-All analyses are conducted in the R programming language and environment [@rcoreteam2022]. For more details on these models and the cost-distance analysis, please see the supplementary materials.
 
@@
 
-# References
+```{r}
+#| label: fig-context-photos
+#| fig-align: center
+#| out-width: 5.5in
+#| fig-cap: "**Context photographs.** (A) Spring San Francisco, an oasis around a lentic pool in a shallow arroyo.
+(B) Abandoned ranch near Spring San Francisco.
+(C) Aerial overview of Arroyo La Presa and a cluster of ranches in the western part of the study.
+The area is criss-crossed by graded and two-track dirt roads, with a lentic pool in the foreground and huertas or small gardens in the background."
 
-::: {#refs}
-:::
+here("figures", "context-photos.png") |> include_graphics()
+
+```
+
+{{< pagebreak >}}
+
+```{r}
+
+table_defaults <- function(x){
+
+  x |> 
+    tab_options(
+      column_labels.font.weight = "bold",
+      data_row.padding = px(5),
+      heading.align = "left",
+      heading.border.bottom.style = "hidden",
+      row_group.as_column = TRUE,
+      table.border.top.style = "hidden",
+      table.border.bottom.style = "hidden"
+    ) |> 
+    sub_missing(missing_text = "")
+
+}
+
+```
+
+```{r}
+#| label: tbl-variables
+#| fig-align: center
+#| out-width: "100%"
+
+fn <- here("_misc", "tbl-variables.png")
+
+here("data", "variable-summary.csv") |> 
+  read_csv() |> 
+  mutate(
+    description = case_when(
+      variable == "paz"        ~ "Time to leave community on way to La Paz (hours)",
+      variable == "cities"     ~ "Time to leave community on way to La Paz from south 
+                                  or Ciudad Constitucion from north (hours)",
+      variable == "springs"    ~ "Mean time to two nearest springs (hours)",
+      variable == "secure"     ~ "Proportion of cluster with clear title or *ejido* 
+                                  membership (0 to 1)",
+      variable == "occupied"   ~ "Cluster has at least one resident (1) or no 
+                                  residents (0)",
+      variable == "population" ~ "Number of individuals residing at ranch cluster"
+    )
+  ) |> 
+  select(model, variable, description, everything()) |> 
+  group_by(model) |> 
+  gt() |> 
+  table_defaults() |>  
+  tab_stubhead("model") |>
+  tab_header(title = "Variable Summary") |> 
+  fmt_number(decimals = 2) |> 
+  fmt_markdown(description) |> 
+  cols_align("left", description) |> 
+  gtsave(filename = fn)
+
+fn |> include_graphics()
+
+```
+
+{{< pagebreak >}}
+
+```{r}
+#| label: tbl-models
+#| fig-align: center
+#| out-width: "80%"
+
+here("data", "model-statistics.rdata") |> load()
+
+fn <- here("_misc", "tbl-models.png")
+
+model_statistics |>
+  gt() |> 
+  table_defaults() |> 
+  tab_header(title = "Summary Statistics") |> 
+  gtsave(filename = fn)
+
+fn |> include_graphics()
+
+```
+
+{{< pagebreak >}}
+
+```{r}
+#| label: tbl-coefficients
+#| fig-align: center
+#| out-width: "65%"
+
+fn <- here("_misc", "tbl-coefficients.png")
+
+coefficient_estimates |>
+  group_by(model) |> 
+  gt() |> 
+  table_defaults() |> 
+  tab_stubhead("model") |> 
+  tab_header(title = "Coefficient Estimates") |> 
+  gtsave(filename = fn)
+
+fn |> include_graphics()
+
+```

### 2024-01-25 R/prepare-images-for-publication.R
trim whitespace around figures

@@ -1,0 +1,32 @@
+
+library(here)
+library(magick)
+library(tidyverse)
+
+# trim all white space around image, 
+# but then add a few white pixels back (dx, dy)
+prepare_image <- function(x, dx = 2, dy = 30, color = "white") {
+  
+  img <- magick::image_read(path = x)
+  
+  img <- magick::image_trim(img)
+  
+  info <- magick::image_info(img)
+  
+  new_width <- info[["width"]] + dx
+  new_height <- info[["height"]] + dy
+  
+  img <- magick::image_extent(
+    img, 
+    geometry_area(new_width, new_height), 
+    color = color
+  )
+  
+  magick::image_write(img, path = x)
+  
+}
+
+walk(
+  list.files(here("figures"), full.names = TRUE), 
+  prepare_image
+)

### 2024-01-25 manuscript/manuscript.qmd
remove multiple bib entries for same reference

@@
 rural ranchers living in the arid Sierra de la Giganta region of Baja California
-Sur (BCS), Mexico [@macfarlan2020a, see @fig-overview-map]. We first explore how
-market integration and local ecology structure the distribution of occupied and
-abandoned ranch clusters. This is a process that took place over several hundred
-years, with the shift to its current distribution beginning in the late 1970s,
-following construction of the transpeninsular Federal Highway 1 [@barkenbus1974;
-@mathes2015]. We then ask how market integration and local ecology have shaped
-the distribution of the population across occupied clusters today.
+Sur (BCS), Mexico [@macfarlan_etal2020b, see @fig-overview-map]. We first
+explore how market integration and local ecology structure the distribution of
+occupied and abandoned ranch clusters. This is a process that took place over
+several hundred years, with the shift to its current distribution beginning in
+the late 1970s, following construction of the transpeninsular Federal Highway 1
+[@barkenbus1974; @mathes2015]. We then ask how market integration and local
+ecology have shaped the distribution of the population across occupied clusters
+today.
 
@@
 provide a small supplement to human and livestock diets during dry periods
-[@macfarlan2020a; @macfarlan_etal2020a; @schniter2021; @lerback2022]. To make
-use of springs, ranchers either have to pump water themselves, using gas or
+[@macfarlan_etal2020b; @macfarlan_etal2020a; @schniter2021; @lerback2022]. To
+make use of springs, ranchers either have to pump water themselves, using gas or
 solar powered equipment, or transport it by gravity irrigation [@lerback2022]. A
@@
 incentives for *ejidos* to parcelize collective land holds and allocate private
-rights to individuals [@barnes2009]. While note the sole cause, this new
-property regime has increased uncertainty and insecurity about land title and
-exposed the region to speculation from real estate markets. As has happened
-elsewhere, these privatization efforts appear to be led by NGOs, commercial
-interests, and other external forces, with the result being greater
-fragmentation of pastoral rangeland systems and increased uncertainty about land
-use [@burnsilver_etal2008; @galvin2009]. For the rural ranchers of the Sierra de
-la Giganta, this process of fragmentation and increasing insecurity at least
+rights to individuals [@barnes2009]. While not the sole cause, this new property
+regime has increased uncertainty and insecurity about land title and exposed the
+region to speculation from real estate markets. As has happened elsewhere, these
+privatization efforts appear to be led by NGOs, commercial interests, and other
+external forces, with the result being greater fragmentation of pastoral
+rangeland systems and increased uncertainty about land use
+[@burnsilver_etal2008; @galvin2009]. For the rural ranchers of the Sierra de la
+Giganta, this process of fragmentation and increasing insecurity at least
 partially explains why ranches of related individuals tend to cluster.
@@
 [@maya1997], with the only permanent sources of naturally occurring freshwater
-being arid-land springs [@grismer1993; @macfarlan2020a; @macfarlan_etal2020a;
-@macfarlan2021]. @lerback2022 provides a fine-grained analysis of these spring
-systems in the project area, describing them as primarily lentic pools at low
-points in the arroyos (see @fig-context-photos B).
+being arid-land springs [@grismer1993; @macfarlan_etal2020b;
+@macfarlan_etal2020a; @macfarlan_etal2021]. @lerback2022 provides a fine-grained
+analysis of these spring systems in the project area, describing them as
+primarily lentic pools at low points in the arroyos (see @fig-context-photos B).
 
@@
 this region in the 18th century, they occupied the largest springs in close
-proximity to arable land [@crosby1994; @macfarlan_etal2020;
+proximity to arable land [@crosby1994; @macfarlan_etal2020b;
 @macfarlan_etal2021]. As the population expanded over time, individuals would

### 2024-01-25 manuscript/manuscript.qmd
change 'binomial' to 'Bernoulli'

@@
 that the dependent variable, whether a ranch cluster is occupied or abandoned,
-has a binomial distribution and is related to the linear predictor by the logit
+has a Bernoulli distribution and is related to the linear predictor by the logit
 link: $$
 \begin{aligned}
-Y &\sim B(n, p) \\
-E[Y/n] &= p \\
+Y &\sim B(1, p) \\
+E[Y] &= p \\
 logit(p) &= \beta_{0} + \beta_{1} D(La\; Paz) + \beta_{2} D(Springs)
 \end{aligned}
-$$ with $n$ being the number of ranch clusters, $p$ the probability that a ranch
-cluster is occupied, and $D(x)$ the travel time from a ranch cluster to $x$ (see
-below for more details about how we calculate this).
+$$ with $p$ being the probability that a ranch cluster is occupied and $D(x)$
+the travel time from a ranch cluster to $x$ (see below for more details about
+how we calculate this).
 

### 2024-02-19 manuscript/manuscript.qmd
additional round of reviews and revisions

@@
 an outsized role in driving these changes, as they introduced powerful new
-incentives for *ejidos* to parcelize collective land holds and allocate private
-rights to individuals [@barnes2009]. While not the sole cause, this new property
-regime has increased uncertainty and insecurity about land title and exposed the
-region to speculation from real estate markets. As has happened elsewhere, these
-privatization efforts appear to be led by NGOs, commercial interests, and other
-external forces, with the result being greater fragmentation of pastoral
-rangeland systems and increased uncertainty about land use
-[@burnsilver_etal2008; @galvin2009]. For the rural ranchers of the Sierra de la
-Giganta, this process of fragmentation and increasing insecurity at least
+incentives for *ejidos* to parcelize collective land holdings and allocate
+private rights to individuals [@barnes2009]. While not the sole cause, this new
+property regime has increased uncertainty and insecurity about land title and
+exposed the region to speculation from real estate markets. As has happened
+elsewhere, these privatization efforts appear to be led by NGOs, commercial
+interests, and other external forces, with the result being greater
+fragmentation of pastoral rangeland systems and increased uncertainty about land
+use [@burnsilver_etal2008; @galvin2009]. For the rural ranchers of the Sierra de
+la Giganta, this process of fragmentation and increasing insecurity at least
 partially explains why ranches of related individuals tend to cluster.
@@
 ecological constraints that characterize this region, especially the aridity of
-the region, we expect water management to play a substantial role in individual
+the region, we expect water resources to play a substantial role in individual
 settlement decisions. Privatization may impose an additional constraint on this
@@
 has a Bernoulli distribution and is related to the linear predictor by the logit
-link: $$
+link:$$
 \begin{aligned}
@@
 \end{aligned}
-$$ with $p$ being the probability that a ranch cluster is occupied and $D(x)$
-the travel time from a ranch cluster to $x$ (see below for more details about
-how we calculate this).
+$$ {#eq-logit-model} with $p$ being the probability that a ranch cluster is
+occupied and $D(x)$ the travel time from a ranch cluster to $x$ (see below for
+more details about how we calculate this).
 
@@
 variable, the population size of a ranch cluster, has a negative binomial
-distribution and is related to the linear predictor by the log link: $$
+distribution and is related to the linear predictor by the log link:$$
 \begin{aligned}
@@
 \end{aligned}
-$$ with $\lambda$ being the average number of individuals per ranch cluster, $k$
-a dispersion parameter, and $P(Security)$ the proportion of ranches in a ranch
-cluster that are secure. The use of a negative binomial rather than the standard
-Poisson model for count data is preferred here to account for dispersion, or
-circumstances in which the variance deviates from the mean, which violates a key
-assumption of the Poisson distribution.
+$$ {#eq-poisson-model} with $\lambda$ being the average number of individuals
+per ranch cluster, $k$ a dispersion parameter, and $P(Security)$ the proportion
+of ranches in a ranch cluster that are secure. The use of a negative binomial
+rather than the standard Poisson model for count data is preferred here to
+account for dispersion, or circumstances in which the variance deviates from the
+mean, which violates a key assumption of the Poisson distribution.
 
@@
 established. We then use travel time to both Ciudad Constitucion and La Paz (ii)
-in the negative binomial GLM examining ranch-cluster population sizes. travel
+in the negative binomial GLM examining ranch-cluster population sizes. Travel
 time to springs (iii) is used in both models, on the assumption that the
@@
 with travel time to cities ($\beta$ = -0.802, *p* \< 0.001), for both
-Constitucion and La Paz, but does not significantly co-vary with travel time to
+Constitucion and La Paz, but is not significantly correlated with travel time to
 springs ($\beta$ = 0.699, *p* = 0.299) (@tbl-coefficients, @fig-response-plots).
@@
 measure of both market integration and local adaptation. Using this simple
-metric has a number of advantages. First, it provides a straightforward way of
-evaluating the relative contribution of each variable to the outcome of
-interest, in this case, settlement decisions among rural ranchers. Second, it
-conforms with geographic theory regarding spatial interactions and dependence
-(otherwise known as Tobler's Law, @tobler1970) and overcomes known limitations
-of Euclidean distance as a cost measure. Finally, it is a relatively simple
-matter to derive those cost estimates from easily accessible remote-sensed
-elevation and road network data held in online repositories. High resolution
-informant data, on the other hand, are generally difficult and costly to
-acquire, often with little chance of providing the right sort of overlap between
-different variables required for fair comparisons.
+metric has a number of advantages. First, it is corroborated by previous
+research [@johansson_etal2002; @corsi_etal2017]. Second, it provides a
+straightforward way of evaluating the relative contribution of each variable to
+the outcome of interest, in this case, settlement decisions among rural
+ranchers. Third, it conforms with geographic theory regarding spatial
+interactions and dependence (otherwise known as Tobler's Law, [@tobler1970]) and
+overcomes known limitations of Euclidean distance as a cost measure. Finally, it
+is a relatively simple matter to derive those cost estimates from easily
+accessible remote-sensed elevation and road network data held in online
+repositories. High resolution informant data, on the other hand, are generally
+difficult and costly to acquire, often with little chance of providing the right
+sort of overlap between different variables required for fair comparisons.
 
@@
 [@henrich2010]. Still, Tobler's Law would seem to suggest that all the variables
-strongly associated with market integration will also co-vary with proximity to
-cities and their markets. In other words, the costs and benefits of market
-integration will positively co-vary with access to markets - both will increase
-as one gains greater access to and interacts more often with markets. So, the
-closer individuals are, the more likely they should be to derive their caloric
-requirements from cities, to work in cities, to own televisions, cellphones, and
-other goods sold in cities, but also the more likely they should be to face
-intense competition for jobs, higher rents, and other costs associated with
-living in that proximity.
+strongly associated with market integration will also be correlated with
+proximity to cities and their markets. In other words, the costs and benefits of
+market integration will be positively correlated with access to markets - both
+will increase as one gains greater access to and interacts more often with
+markets. So, the closer individuals are, the more likely they should be to
+derive their caloric requirements from cities, to work in cities, to own
+televisions, cellphones, and other goods sold in cities, but also the more
+likely they should be to face intense competition for jobs, higher rents, and
+other costs associated with living in that proximity.
 
@@
 than being an outcome from which to draw conclusions, we believe this result
-should be interpreted with caution. Our measure of land security is based on a
-combination of government data and informant observations and so likely does not
-capture the full complexity of this important issue. A larger dataset with a
-more standardized and quantifiable measure of land security would be more
-representative. We also note that the issue of land security is not equally
-salient to everyone in this population. Many of the informants that might be
-land insecure by most objective measures do not necessarily act as if their
-title to land is uncertain. This would suggest that perceptions of risk with
-respect to property rights should be studied more thoroughly.
+should be interpreted with caution as previous scholarship suggests that land
+tenure is essential for securing the livelihood of smallholders
+[@corsi_etal2017; @besley_ghatak2010; @wickeri_kalhan2010;
+@aggarwal_freudenberger2013; @dejanvry_etal2015] and ensuring sustainable
+resource management. Furthermore, our research on the effect of land security on
+ranching decision-making in this population suggests that it influences whether
+or not households engage in cattle production -- a risky economic decision in
+this arid ecosystem [@macfarlan_etal2023a]. So, what explains our negative
+result? Our argument is threefold. First, the distribution of land secure and
+insecure households is a result of changes to the property regime that happened
+in the 1990s [@dejanvry_etal2015], around the time when the distribution of
+ranches and ranching families was established, so there may be a lag effect that
+our analysis is insensitive to. With additional data on the distribution of
+households across a larger spatial area (or temporal window, for the
+establishment of new households into the future), we may find that land
+insecurity is having a real effect on individual choices. Second, our measure of
+land security is based on a combination of government data and informant
+observations and so likely does not capture the full complexity of this
+important issue. A larger dataset with a more standardized and quantifiable
+measure of land security would be more representative. Finally, we note that the
+issue of land security, particularly as that is understood as clear title, is
+not equally salient to everyone in this population [@lambrecht_asare2016]. Many
+of the informants that might be land insecure by most objective measures do not
+necessarily act as if their title to land is uncertain. This would suggest that
+perceptions of risk with respect to property rights should be studied more
+thoroughly.
 
@@
 within a largely agrarian economy [@shenk2016]. Currently, the ranching
-communities of BCS are characterized by more intensive kin networks, but the
-limited available evidence suggests that this might be changing, with
-individuals increasingly marrying outside of the ranching community, most likely
-to individuals living within the cities. Families are as a consequence
-increasing their connections to the cities and by extension increasing the
-incentives for integrating more closely with their markets.
+communities of BCS are characterized by more intensive kin networks
+[@macfarlan_etal2020b], but the limited available evidence suggests that this
+might be changing, with individuals increasingly marrying outside of the
+ranching community, most likely to individuals living within the cities
+[@page_etal2024]. Families are as a consequence increasing their connections to
+the cities and by extension increasing the incentives for integrating more
+closely with their markets.
 
@@
 ```{r}
-#| label: tbl-models
-#| fig-align: center
-#| out-width: "80%"
-
-here("data", "model-statistics.rdata") |> load()
-
-fn <- here("_misc", "tbl-models.png")
-
-model_statistics |>
-  gt() |> 
-  table_defaults() |> 
-  tab_header(title = "Summary Statistics") |> 
-  gtsave(filename = fn)
-
-fn |> include_graphics()
-
-```
-
-{{< pagebreak >}}
-
-```{r}
 #| label: tbl-coefficients
@@
 #| out-width: "65%"
+
+here("data", "model-statistics.rdata") |> load()
 
@@
 ```
+
+{{< pagebreak >}}
+
+```{r}
+#| label: tbl-models
+#| fig-align: center
+#| out-width: "80%"
+
+fn <- here("_misc", "tbl-models.png")
+
+model_statistics |>
+  gt() |> 
+  table_defaults() |> 
+  tab_header(title = "Summary Statistics") |> 
+  gtsave(filename = fn)
+
+fn |> include_graphics()
+
+```

### 2024-02-26 R/least-cost-paths.qmd
add 80 character linebreaks to qmd

@@
 
+```{=html}
 <!-- 
@@
 -->
-
+```
 ```{css}
@@
 
-**Goal:** estimate cost-distance (in time) from each ranch to Ciudad Constitucion, La Paz, and local springs.
+**Goal:** estimate cost-distance (in time) from each ranch to Ciudad
+Constitucion, La Paz, and local springs.
 
-**Data:** a Digital Elevation Model (DEM) along with point locations for ranche clusters, springs, cities.
+**Data:** a Digital Elevation Model (DEM) along with point locations for ranche
+clusters, springs, cities.
 
@@
 
-There are two watersheds in the project area.
-Each one has a single road that connects it to Highway 1 and thus to the markets in Ciudad Constitucion to the north and the capitol La Paz to the south.
-The "road" that connects the watersheds to each other is not maintained and thus rarely used.
-As a consequence, nearly all of the variance in travel time to cities is found within each watershed.
-For that reason, we estimate the time it takes to leave each watershed ("leaving" here means hitting the edge of the project area), rather than the true time it takes to get to each city.
-We interpret this as the relative difference in time it takes to get to Ciudad Constitucion from the northern watershed and the relative difference in time it takes to get to La Paz from the southern watershed.
-To get to La Paz from the northern watershed, we simply add an arbitrary half hour to the total time to get out of that watershed.
-The same would presumably be the case to get to Ciudad Constitucion from the southern watershed, though that value does not factor into our analysis.
+There are two watersheds in the project area. Each one has a single road that
+connects it to Highway 1 and thus to the markets in Ciudad Constitucion to the
+north and the capitol La Paz to the south. The "road" that connects the
+watersheds to each other is not maintained and thus rarely used. As a
+consequence, nearly all of the variance in travel time to cities is found within
+each watershed. For that reason, we estimate the time it takes to leave each
+watershed ("leaving" here means hitting the edge of the project area), rather
+than the true time it takes to get to each city. We interpret this as the
+relative difference in time it takes to get to Ciudad Constitucion from the
+northern watershed and the relative difference in time it takes to get to La Paz
+from the southern watershed. To get to La Paz from the northern watershed, we
+simply add an arbitrary half hour to the total time to get out of that
+watershed. The same would presumably be the case to get to Ciudad Constitucion
+from the southern watershed, though that value does not factor into our
+analysis.
 
-For springs, Dr.
-MacFarlan's ethnographic data suggest that individuals residing within the same ranch cluster tend to rely on one or two springs at most.
-Thus, we estimate road distance to the two nearest springs and take the average.
+For springs, Dr. MacFarlan's ethnographic data suggest that individuals residing
+within the same ranch cluster tend to rely on one or two springs at most. Thus,
+we estimate road distance to the two nearest springs and take the average.
 
-All spatial data are projected to the Mexico ITRF92 / UTM zone 12N CRS ([EPSG:4485](epsg.io/4485)).
+All spatial data are projected to the Mexico ITRF92 / UTM zone 12N CRS
+([EPSG:4485](epsg.io/4485)).
 
@@
 
-:::{.mx-auto .my-4 .img-45}
+::: {.mx-auto .my-4 .img-45}
 ![](../figures/od-points.png)
@@
 
-To estimate least-cost paths, we first derive slope estimates from the DEM then calculate the hiking speed at which individuals can traverse any given cell using Campbell's hiking function.
-After that, we decrease the cost in units of time to traverse grid cells crossed by roads and other paths using weights that approximate travel times by car or horse.
-This is equivalent to increasing the speed one travels in those cells.
-Below are the specific weights we use.
+To estimate least-cost paths, we first derive slope estimates from the DEM then
+calculate the hiking speed at which individuals can traverse any given cell
+using Campbell's hiking function. After that, we decrease the cost in units of
+time to traverse grid cells crossed by roads and other paths using weights that
+approximate travel times by car or horse. This is equivalent to increasing the
+speed one travels in those cells. Below are the specific weights we use.
 
-:::{.w-50 .mx-auto .py-4}
+::: {.w-50 .mx-auto .py-4}
 | Type      | Weight | Approx. Speed     |
@@
 
-:::{.mx-auto .my-4 .img-45}
+::: {.mx-auto .my-4 .img-45}
 ![](../figures/cost-surface.png)
@@
 
-Please note that travel time is not strictly to cities, but to the edge of the project area.
-As noted above, for the northern watershed, the time to La Paz equals the time it takes to leave the watershed plus half an hour.
+Please note that travel time is not strictly to cities, but to the edge of the
+project area. As noted above, for the northern watershed, the time to La Paz
+equals the time it takes to leave the watershed plus half an hour.
 
@@
 
-:::{.mx-auto .my-4 .img-65}
+::: {.mx-auto .my-4 .img-65}
 ![](../figures/least-cost-paths.png)
